---
title: "Final Project Work"
author: "Caroline Gheen"
date: "2026-04-21"
output:
  word_document:
    fig_height: 6
    fig_width: 8
  pdf_document: default
  html_document:
    fig_height: 6
    fig_width: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning = FALSE, include = FALSE}
# load packages
library(dplyr) # for piping
library(car) # for recoding
library(usmap) # for heatmap
library(ggplot2) # for graphing
source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt") # source for pairsJDRS and MyResPlots
```

# Introduction
I utilized the American Hospital Association's Annual Survey IT supplement as my data source, specifically the 2024 answers. This is an annual survey sent out to all hospitals that are member of the American Hospital Association and asseses their technology and information technology status. It is required and typically filled out by an institutions' CIO.
I am studying Health Informatics at YSPH, so I am very interested in the information systems at hospitals and this is a dataset I have been wanting to explore. My analysis mostly hinges on Health Information Exchanges (HIE) which are networks in which hospitals participate to share information with each other, mainly to improve public health or patient care. 

```{r, include = FALSE}
# load initial data
data <- read.csv("AHA Data - Query 2.csv")
dim(data) # 2641 institutions responded, only loaded 48 variables; in total 27 questions are asked
names(data)
str(data)
```

# DATA 

I isolated 4 questions: 
"Q-10: Indicate level of participation in state, regional, and/or local health information exchange (HIE) or health information organization (HIE),"
Options: HIE/HIO is operational in my area and we are participating and actively exchanging data in at least one HIE, HIE/HIO is operational in my area but we are not participating, HIE/HIO is not operational in my area, do not know

"Q-11: Which of the following national health information exchange networks and/or frameworks does your hospital currently participate in,"
Options: CommonWell Health Alliance, e-Health Exchange, Civitas, EHR vendor network, Other national HIE and describe, do not participate in, do not know

"Q-14: How does your hospital submit data for public health reporting related to activities below," 
Options: Syndromic surveillance reporting, Immunization registry reporting, Electronic case reporting, public health registry reporting, clinical data registry reporting, electronic reportable laboratory result reporting, antimicrobial use and resistance, hospital capavity and utilization of medical records

"Q-18: Does your hospital use any machine learning or other predictive models that display output or recommendations"
Options: Yes or No to machine learning, other and describe 

"Q-19: To which of the following uses has your hospitals applied machine learning or other predictive models"
Options: Yes or No to predicting health trajectories or risk for inpatients, identify high risk options for follow up, monitor health, recommend treatments, simplify or automate billing, faciliate scheduling, other and describe (operational process optimization), other and describe (clinical use cases)

Ownership: government non-federal (state, county, city, city-county, hospital district or authority), non-government/not-for-profit (church operated, other), investor owned/for-profit (individual, partnership, corporation), government/federal (Department of Defense, public health service, Veterans Affairs, federal other, Indian service, Department of Justice)

and 4 descriptors: AHA ID number, hospital name, total facility beds set up and staffed, and primary Electronic Medical Record (EMR) used.

Question 11, 14, 18, and 19 are stored as individual variables for each options and thus are similar to binary indicators in that if that option is selected it is a 1 and if it was not, it is a 0. Because of this, for question 14, I only exported the variables that are each type of reporting into the HIE because I am only concerned with if institutions use their HIE for that mandaotry reporting.
There are 4 'describe other' options which are free text. These are if you use an other primary EMR, active in another HIE,use of another predictive models, use for ML or predictive that is not listed. 

I created two composite variables: composite HIE score and AI score. To create composite HIE score, I added 1 if they participate in an HIE (question 10), added 1 for each HIE they participate in (question 11) and added 1 for each public health reporting in which they utilized HIE (question 14).
For AI score, I added 1 for if they use ML or another predictive model (question 18) and added 1 for each activity they apply AI for (question 19).

Ultimately these were the variables I was most concerned with:
1. HIE Participation Score (continuous out of 14)
2. AI Composite Score (continuous out of 10)
3. Number of Beds (continuous)
4. Ownership type (categorical)
5. Primary EMR (categorical)
6. State (categorical)

# Data Cleaning and Descriptive Plots
The first thing I did was change the names of the variable to make them more human readable and help track which question each variable came from. Although I'm concerned with only 5 overall variables, my dataset contained about 48 variables. 

```{r, include = FALSE}
# change names to be more human readable
names(data)[names(data) == "piemro"] <- "primary_emr_other"
names(data)[names(data) == "mcntrl"] <- "ownership"
names(data)[names(data) == "piemr"] <- "primary_emr"
names(data)[names(data) == "rhio_lvl1"] <- "q10_HIE_partic"
names(data)[names(data) == "othnet_desc"] <- "q11_other_nat_desc"
names(data)[names(data) == "cwha"] <- "q11_commonwell"
names(data)[names(data) == "hlthxch"] <- "q11_ehealth"
names(data)[names(data) == "netdnk"] <- "q11_dnk_othernat"
names(data)[names(data) == "netno"] <- "q11_no_nat"
names(data)[names(data) == "othnet"] <- "q11_other_nat"
names(data)[names(data) == "shiec"] <- "q11_civitas"
names(data)[names(data) == "spcareq"] <- "q11_carequality"
names(data)[names(data) == "vennet"] <- "q11_ehrbased"
names(data)[names(data) == "ssrstrdt"] <- "q15_syndromic"
names(data)[names(data) == "irrstrdt"] <- "q15_immune"
names(data)[names(data) == "ecrstrdt"] <- "q15_electronic"
names(data)[names(data) == "phrstrdt"] <- "q15_phreg"
names(data)[names(data) == "cdrstrdt"] <- "q15_clinicreg"
names(data)[names(data) == "erlstrdt"] <- "q15_elab"
names(data)[names(data) == "hcustrdt"] <- "q15_hospcap"
names(data)[names(data) == "ssrhie"] <- "q14_syndromic_hie"
names(data)[names(data) == "iirhie"] <- "q14_immune_hie"
names(data)[names(data) == "ecrhie"] <- "q14_electronic_hie"
names(data)[names(data) == "phrhie"] <- "q14_phreg_hie"
names(data)[names(data) == "cdrhie"] <- "q14_clinicreg_hie"
names(data)[names(data) == "erlhie"] <- "q14_elab_hie"
names(data)[names(data) == "hcuhie"] <- "q14_hospcap_hie"
names(data)[names(data) == "amichie"] <- "q14_amr_hie"
names(data)[names(data) == "aitraj"] <- "q19_ai_traj"
names(data)[names(data) == "airfol"] <- "q19_ai_risk_fu"
names(data)[names(data) == "aimhea"] <- "q19_ai_monitor"
names(data)[names(data) == "airect"] <- "q19_ai_rec"
names(data)[names(data) == "aibill"] <- "q19_ai_bill"
names(data)[names(data) == "aische"] <- "q19_ai_schedule"
names(data)[names(data) == "aipoth"] <- "q19_ai_process"
names(data)[names(data) == "aipoth_desc"] <- "q19_ai_process_desc"
names(data)[names(data) == "aicloth"] <- "q19_ai_clinic"
names(data)[names(data) == "aicloth_desc"] <- "q19_ai_clinic_desc"
names(data)[names(data) == "ainota"] <- "q19_ai_none"
names(data)[names(data) == "aidnk"] <- "q19_ai_dnk"
names(data)[names(data) == "aurstrdt"] <- "q15_amr"
names(data)[names(data) == "predml"] <- "q18_ml"
names(data)[names(data) == "predoth"] <- "q18_otherml"
data$othpad_desc <- NULL
```

```{r, echo = FALSE}
names(data)
```

The first thing I did was explore missingness. To do this, I made a dataframe that was each variable, the amount of data present, amount missing, and the percent complete. Overall, there was low missingness, so I questioned if a lot of the missingness was potentially coming from the same institutions just not turning in thorough surveys. To do this, I made another dataframe that was each instiutions ID, the amount of data present, amount of missing, and percent complete. 

```{r, include = FALSE}
# make df so easily see for each column how many missing, complete and % complete 
na_summary <- data.frame(
  n_data = colSums(!is.na(data)),
  n_na = colSums(is.na(data)),
  pct_comp = round(colSums(!is.na(data)) / nrow(data) * 100, 1)
)

# there is lots of really good data - not a lot of missingness for columns
# I wonder if the missingness to the columns  are coming from the same organizations ie some were really thorough in filling out the survey and some were not

# look at rows- if same organizations are NA for all, will potentially drop them
row_na_summary <- data.frame(
  id = data$id,
  n_data = rowSums(!is.na(data)),
  n_na = rowSums(is.na(data)),
  pct_comp = round(rowSums(!is.na(data)) / ncol(data) * 100, 1)
)
```

```{r,include = FALSE}
## include table
names(na_summary) <- c("Complete_Data", "Missing_Data", "Percent_Complete")
names(row_na_summary) <- c("ID", "Complete_Data", "Missing_Data", "Percent_Complete")

head(na_summary)
tail(na_summary)

head(row_na_summary)
```

Looking at the first few row entries, a lot of organizations have 100% complete data, some have 90-100%, even less have 60-90%, and some have around 20%. I definitely want to drop the ones that only have 20% of data complete, but I wanted to explore how "harsh" can I be while still retaining a meaningful number of entries. To do this, I did three subsets of keeping all entries that have a percent complete over 61%, over 90%, and 100% complete. With 100% complete, this drops 1128 institutions but retains 1513 institutions. I am happy with this sample size and it gives me fully complete data, so I made this data2 that becomes the official dataset used throughout the rest of the project. 
```{r, include = FALSE}
# just looking a the first few row entries, a lot of organizations have 100% complete, some have 90-100%, even less have 60-90%, and some have around 20%. I definitely want to drop the ones that only have 20% of data complete, but how harsh can I be while still retaining a meaningful number of entries?
# what is my nrow difference (ie how many institutions I retain) if I make a harsh boundary of only keep those who completed 90%+ vs those who completed 60%+. Or even if I only kept those who answered all questions
nrow(row_na_summary[row_na_summary$Percent_Complete > 61, ]) # 2485
nrow(row_na_summary[row_na_summary$Percent_Complete > 90, ]) # 2098
nrow(row_na_summary[row_na_summary$Percent_Complete == 100.0, ]) # 1513

# select selects that are full as best
best_rows <-(row_na_summary[row_na_summary$Percent_Complete == 100.0, ])

# keep only data that has all rows full as data2
data2 <- data %>% filter(id %in% best_rows$ID) 

# sanity check that there are no NA - there are empty values but these are meaningful (ie. no primary other emr means they don't use other primary emr)
sum(is.na(data2)) # sanity check - 0
```

```{r, include = FALSE}
# we dropped 1128 institutions off of missingness. This means we  can do full comparisons (don't have to use pairwise comparisons, for example). We retained 1513 institutions
nrow(data) - nrow(data2) #1128
nrow(data2) #1513
```

Next, all questions have No keyed as 0 except question 18 has No keyed as 2. So, I would like to recode so No is consistantly 0 throughout the entire data.
```{r, include = FALSE}
# recode question 14
data2$q18_ml <- recode(data2$q18_ml, "2 = 0; 3 = 2" )

data2$q18_otherml <- recode(data2$q18_otherml, "2 = 0; 3 = 2" )

unique(data2$q18_ml)
```

```{r, include = FALSE}
str(data2)
# structure seems good for all. my coding made them numeric instead of integer - this is probably fine. will likely need to make these factor to do categorical things. can come back and do this once I decide
```

Next, I want to look at if all states are represented. There are 52 unique state entries with all 50 states being represented plus two territories. One territory is PR which is Puerto Rico and one is MP which I am unfamiliar with. By filtering to only this entry and Googling the institution name, I discover it is Mariana Islands in the Pacific. 
```{r, include = FALSE}
length(unique(data2$mstate)) # 52 states - PR and MP in addition to all 50 states

# I have no idea what MP is...
data2[data2$mstate == "MP",]
# googling the name of the health center and MP I discover it is in Mariana Islands for Navy members stationed there
```

```{r, include = FALSE}
# okay so we know all 50 state plus some are represented - how many hospitals are in each
table(data2$mstate)
# make this into a df so it is plottable
state_counts <- as.data.frame(table(data$mstate))
colnames(state_counts) <- c("state", "n")

state_counts %>% arrange(desc(n))
# Texas has 266 hospitals, but the mean of hospitals per state is 49... the minimum is 1 hospital in a state (Mariana Islands)
summary(state_counts$n)

barplot(state_counts$n)
boxplot(state_counts$n)
hist(state_counts$n) # this is right skewed as is expected
# it would be cool to come back and do some sort of frequency plot that makes the state name bigger the more hospitals it has
```

Overall, the amount of hospitals distributed per states varies - Texas has 266 hospitals while Mariana Pacific only has 1.
```{r, echo = FALSE}
## include about states 
summary(state_counts$n)

hist(state_counts$n,
     main = "Histogram of Hospitals per State",
     xlab = "Amount of Hospitals per State",
     ylim = c(0, 35),
     col = "tomato2")
```

Next, looking at the type of primary EMRs that are used. Cerner (owned by Oracle), Epic, and Meditech are the giants in this industry and have the most users as expected. 
```{r, include = FALSE}
table(data2$primary_emr)
# lots at Cerner/Oracle(2) and Epic (5) and Meditech(9) as expected. Also lots ie 50+ at Other(91 - these are in the other variable), CPSI/Evident(17).

as.data.frame(table(data2$primary_emr_other))
# Wellsky has a lot of responses (32)
# MyAvatar that is also NetSmart has lots of responses but they aren't going together based on different spelling and capitalization - I don't think it will add together high enough to become it's onw category so i don't think i'll spend the time to standardize

length(unique(data2$primary_emr_other)) # only 37 specified the other system they used but 91 said they use another system...
```

I am going to condense the options to only be these three and all others will become an 'other' category. (Side note: no institutions reported that they did not have an EMR - this is great and expected as there are large government incentives and penalies for not using an EMR).
```{r, echo = FALSE}
## include 
hist(data2$primary_emr, 
     main = "What Primary EMRs Are Used",
     xlab = "Coding of Primary EMR",
     ylim = c(0, 1000),
     col = "skyblue1")
```


```{r, include = FALSE}
# condense primary emr answers
data2$primary_emr[!data2$primary_emr %in% c(2, 5, 9, 30)] <- 0

# check only 2, 5, 9, and 30 are options for emr
unique(data2$primary_emr)

# new table with condensed options
table(data2$primary_emr)
```

This barplot shows the new distribution and percentage of institutions using that EMR. We see that Epic is still the runaway leader with almost half of the institutions using their Epic. Meditech, Cerner, and other EMRs are used by a similar number of institutions. 

```{r, echo = FALSE}
## include 
primary_emr <- table(data2$primary_emr)
tabpct <- paste0(round(primary_emr / sum(primary_emr) * 100, 0), "%")
names(primary_emr) <- c("Other", "Cerner", "Epic", "Meditech")

barplot(primary_emr, 
        main = "What EMRs are Used (Condensed)",
        xlab = "EMRs",
        col = "skyblue3", 
        ylim = c(0, 1000))
text(c(.75, 1.97, 3.15, 4.35), c(350, 400, 920, 280), tabpct, cex = .8, col = "seashell4")
```

Next, the ownership was originally coded as about 45 different options under 4 broad categories: government, non-federal; not-for-profit; for-profit; government, federal. I condensed the options to be the 4 general categories. 
Not-for-profits are the largest category - this makes sense as it includes the religious institutions and, I believe, the academic medical centers. Federal is smaller than expected; there is only 1 owned by the DOD (this should include military hospitals) and only 9 VAs, which is thought to be the largest healthcare institution, so this is not representative of the actual hospital breakdown and potentially due to only keeping complete responses. 
```{r, include = FALSE}
# look at ownership as is
table(data2$ownership)
table(data$ownership)
# 12-16 is all types of government-nonfederal; 21 and 23 are non-government/not for profits; 31-33 are for profits; 40-48 are federal goverment types 
# not sure why an ownership option is the department of justice but none of those appear in this dataset - maybe if they go bankrupt 

# I am going to combine the individual breakdowns into the larger general groups given in the dataset
# government non-federal = 1; nongov, not for profit = 2; for profit = 3; federal = 4
data2$ownership[data2$ownership %in% c(12, 13, 14, 15, 16)] <- 1

data2$ownership[data2$ownership %in% c(21, 23)] <- 2

data2$ownership[data2$ownership %in% c(31, 32, 33)] <- 3

data2$ownership[data2$ownership %in% c(40, 44, 45, 46, 47, 48)] <- 4

# check current unique values 
unique(data2$ownership)
```

```{r, echo = FALSE}
## include 
ownership <- table(data2$ownership)
tabpct <- paste0(round(ownership / sum(ownership) * 100, 0), "%")
names(ownership) <- c("Gov. non-federal", "Not-for-profit", "For-profit", "Gov. federal")

barplot(ownership,
        xlab= "Owners",
        main = "Ownership Types",
        ylim = c(0, 1300),
        cex.names = .6,
        col = "peachpuff3")
text(c(.75, 1.97, 3.15, 4.35), c(400, 1200, 250, 150), tabpct, cex = .8, col = "seashell4")
```

Bed Total was retained as a proxy for hospital size. The range is 3 to 2919 so a very wide spread. As is, this variable is heavily right skewed, therefore, I applied a log transformation, which will be used throughout the rest of the analysis. Because no values were non-zero I did not need to apply a constant first. 
```{r, include = FALSE}
range(data2$bdtot)
```

```{r, echo = FALSE}
# scatterplot of bed total
# most likely will want to take log transformation of bed total
par(mfrow) = c(1, 2)
hist(data2$bdtot, 
     main = "Total Number of Beds", 
     xlab = "Number of Beds",
     col = "mistyrose")
hist(log(data2$bdtot),
     main = "Log Total Number of Beds",
     xlab = "Log Number of Beds",
     col = "mistyrose3")
data2$logbed <- log(data2$bdtot)
```

I want to explore the free answer text variables: other HIE, other process use of AI, other clinical use of AI.
```{r, include = FALSE}
# 3 free text variables
as.data.frame(table(data2$q11_other_nat_desc))

as.data.frame(table(data2$q19_ai_process_desc))

as.data.frame(table(data2$q19_ai_clinic_desc))
```

```{r, include = FALSE}
data2$q11_other_nat_desc[data2$q11_other_nat_desc == ""] <- NA
data2$q19_ai_clinic_desc[data2$q19_ai_clinic_desc == ""] <- NA
data2$q19_ai_process_desc[data2$q19_ai_process_desc == ""] <- NA

# recheck unique
#unique(data2$q11_other_nat_desc)
#unique(data2$q19_ai_clinic_desc)
#unique(data2$q19_ai_process_desc)
# all "" changed to NA successfully
```

First, I wanted to see if any answers reported in the 'other' free text are actually other coded options that should have been selected instead. 
There were two hospitals who marked they were participating in CommonWell and answered they participated in 'other,' specifying a subgroup of CommonWell, so I recoded their 'other' participation as 0 in order to only count CommonWell once. There were 6 institutions (all from the same parent institution, so likely filled out by the same individual) that marked they participate in an EHR based HIE and specified in 'other' that they participate in Epic, which is an EHR based HIE. I recoded their 'other' participation to 0 so their Epic participation was only counted once. 3 institutions reported they participate in TEFCA, which is an agreement, not an exchange, so I recoded their 'other' to 0. 1 institution reported they participate in eCQM which is a CMS quality measure not an exchange, so I recoded their 'other' to 0. 1 institution reported they participate in National Research Institute which is the sharing of genomes, but I don't believe it counts as an HIE, so I recoded their 'other' to 0. 
```{r, include = FALSE}
# show full entries for any hospital that has an entry in other please specify 
data2[!is.na(data2$q11_other_nat_desc), ]

# other HIE
as.data.frame(table(data2$q11_other_nat_desc))
# 1 for Commonwell (CommonWell through Healthix) - they also marked commonwell so this will just become 0 and other descriptor will become NA
data2[which(data2$q11_other_nat_desc == "CommonWell through Healthix"), ] 
data2$q11_other_nat[which(data2$q11_other_nat_desc == "CommonWell through Healthix")] <- 0 

#6 for Epic - they also maked that they use an ehr based HIE, so this will just become 0 and other descriptor will become NA; all hospitals who said this are Carilion meaning this survey was likely filled out by the same person for mulitple institutions
data2[which(data2$q11_other_nat_desc == "Epic"), ]
data2$q11_other_nat[which(data2$q11_other_nat_desc == "Epic")] <- 0

# 3 for TEFCA - remove; TEFCA is an agreement not a framework; 1 eCQM - remove; eCQM is a CMS quality measure; 1 National Research Institute - NRI does not have an HIE (they share genomes but)- these will become I do not know
data2$q11_other_nat[which(data2$q11_other_nat_desc == "TEFCA")] <- 0
data2$q11_other_nat[which(data2$q11_other_nat_desc == "eCQM")] <- 0
data2$q11_other_nat[which(data2$q11_other_nat_desc == "National Research Institue")] <- 0

```
I also analyzed if any of the 'other' answers could be their own group. Many of the 'other' answers are state exchanges (CA, CO, NJ, CT/MA, KY, IN KA, OH, OK) and federal exchanges. Surescripts, CyncHealth, and Direct Trust also have a sufficient number of answers, but not enough to justify being their own category.

The AI clinical and operational process uses also have a free text box to describe any other use. I noticed a lot of overlap in responses (ex: some institutions reported ambient scribes that help doctors take notes as clinical use and some report it as process uses), leading me to believe there was not a lot of guidance from the AHA in how processes should be categorized. Ultimately, to assign an AI score, I gave +1 if institutions marked 'other,' which is the same as if they had marked another category. The AI score was just if they used AI for that category not how many tools they had. Therefore, it did not affect the scoring if their answer was in the 'other' category or another category. So, all other answers stayed in the other category, although it should be noted that some institutions may use more AI because they have more tools for one particular task rather than for multiple tasks. 
```{r, include = FALSE}
as.data.frame(table(sort(data2$q19_ai_clinic_desc)))
```

```{r, include = FALSE}
as.data.frame(table(sort(data2$q19_ai_process_desc)))
```

Looking deeper into the breakdown of the other questions:
```{r, include = FALSE}
# quesiton 11 - includes all except other frequency
q11df <- data2[, c("q11_carequality", "q11_commonwell", "q11_ehealth", "q11_dnk_othernat", "q11_no_nat", "q11_other_nat", "q11_civitas", "q11_ehrbased")]
q11_freq <- as.data.frame(colSums(q11df))
names(q11_freq) <- c("n")
pct <- paste0(round((q11_freq$n/nrow(data2)*100), 0), "%")
q11_freq <- cbind(q11_freq, pct)
#  % of hospitals a part of that hie 

# question 19 
q19df <- data2[, c("q19_ai_traj","q19_ai_risk_fu", "q19_ai_monitor", "q19_ai_rec", "q19_ai_bill", "q19_ai_schedule", "q19_ai_process", "q19_ai_clinic", "q19_ai_none", "q19_ai_dnk")]
q19_freq <- as.data.frame(colSums(q19df))
names(q19_freq) <- c("n")
pct <- paste0(round((q19_freq$n/nrow(data2)*100), 0), "%")
q19_freq <- cbind(q19_freq, pct)
#  % of hospitals who use that tool 

# question 14
q14df <- data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie")]
q14_freq <- as.data.frame(colSums(q14df))
names(q14_freq) <- c("n")
pct <- paste0(round((q14_freq$n/nrow(data2)*100), 0), "%")
q14_freq <- cbind(q14_freq, pct)
#  % of hospitals who use hie for that ph reporting

```

% of hospitals that participate in specific HIE
```{r, echo = FALSE}
#  % of hospitals that use that hie
rownames(q11_freq) <- c("CareQuality", "CommonWell", "e-Health Exchange", "Do Not Know", "No HIE", "Other", "Civitas", "EHR-based")
par(mar = c(5, 8, 4, 4))
barplot(q11_freq$n,
        main = "Percent Hospitals that Participate in Each HIE",
        xlab = "HIE Groups",
        names = rownames(q11_freq),
        cex.names= .7,
        horiz = TRUE,
        las = 1,
        col = "navajowhite",
        xlim = c(0, 1000))
text(c(950, 80, 275, 175, 125, 650, 630, 880), c(9, 8, 6.7, 5.5, 4.5, 3.2, 2, .75), rev(q11_freq$pct), cex = .6, col = "seashell4", font = 2)
q11_freq
```

% of hospitals who use applied ML or predictive analysis for that task 
```{r, echo = FALSE}
#  % of hospitals who use that tool 
rownames(q19_freq) <- c("Predicting Health Trajectors or Risk for Inpatients", "Identify High Risk Outpatients", "Monitor Health", "Recommend Treatments", "Simplify or Automate Billing", "Facilitate Scheduling", "Other Operation Process", "Other Clinical Uses", "No ML Use", "Do Not Know")
q19_freq
```

```{r, echo = FALSE}
# % oh hospitals who use that hie for ph reporting
rownames(q14_freq) <- c("Syndromic Surveillance", "Immunization Registry", "Electronic Case Reporting", "Public Health Registry", "Clinical Data Registry Reporting", "Electronic Reportable Lab Reporting", "Hospital Capacity and Utilization of Medical Supplies", "Antimicrobial Use and Resistances")
q14_freq
```

To build the AI composite score, I +1 for each instance institutions use ML or predictive models and for how many tasks they use it for. 
```{r, include = FALSE}
# create AI variable off questions 18 and 19
data2$ai_score <- rowSums(
  cbind(
    data2[, c("q19_ai_traj", "q19_ai_risk_fu", "q19_ai_monitor",
              "q19_ai_rec", "q19_ai_bill", "q19_ai_schedule",
              "q19_ai_process", "q19_ai_clinic")],
    ifelse(data2$q18_ml %in% c(0,1), data2$q18_ml, 0),
    ifelse(data2$q18_otherml %in% c(0,1), data2$q18_otherml, 0)
  )
)

range(data2$ai_score) # someone uses AI for all: 0-10
table(data2$ai_score) # a lot of hospitals are still not using AI, but most are integrating multiple uses
barplot(table(data2$ai_score))

table(data2$q18_ml)
table(data2$q18_otherml)
#1090 should be using AI in some way according to question 19 (number of hospitals - those who answered 0 = 1090) but according to question 18, only 800-900 answered yes to those. which do i trust.
```
The majority of institutions are not using AI, but who are using AI in all possible categories.
```{r, echo = FALSE}
## include 
hist(data2$ai_score, 
     main = "Frequency of AI Scores",
     xlab = "AI Scores",
     col = "lightsteelblue",
     ylim = c(0, 500))
```

To score hospitals on their HIE participation and create the HIE participation variable, institutions will get +1 for each public health reporting they utilize HIE for, captured in question 14. Because these are already scored as 0 for no and 1 for yes, we can row sum these variables. Then, they will get +1 for each HIE they participate in. This would mean they have a 1 for any question 11, except do not know "q11_dnk_othernat" or "q11_no_nat"
```{r, include = FALSE}
data2$hie_participation <- rowSums(data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie", "q11_other_nat", "q11_commonwell", "q11_ehealth", "q11_civitas", "q11_carequality","q11_ehrbased")])
```

```{r, include = FALSE}
# create own variable to cut down on typing
hie_part <- data2$hie_participation
range(hie_part) # 0-13; max would have been 14, so nobody uses everything
table(hie_part)
```
We see a lot of institutions do not participate in HIE at all, and no institutions use HIE for all possible public health reporting manners.
```{r, echo = FALSE}
## include 
summary(hie_part)

hist(hie_part, 
     main = "Distribution of HIE Participation",
     xlab = "HIE Participation",
     col = "navajowhite",
     breaks = 14,
     xaxt = 'n',
     ylim = c(0, 500))
axis(1, at = 0:14, labels = 0:14, cex.axis = .8)

```
We will notice this is right skewed. 


Overall HIE Participation:
```{r, echo = FALSE}
hie <- as.data.frame(table(data2$q10_HIE_partic))
hie$pct <- paste0(round(hie$Freq / sum(hie$Freq) * 100, 0), "%")
# the majority of hospitals do participate in an HIE
names(hie) <- c("HIE Status", "Frequency", "Percent")
rownames(hie) <- c("Area HIE is operational and do participate", "Area HIE is operational but do not participation", "Area HIE is not operational", "Do Not Know")

hie
```



Now, if we compare the variables to each other,
Looking at HIE Participation by Bed Total, there does not seem to be any relationship. 
```{r, echo = FALSE}
plot(hie_part ~ log(data2$bdtot), 
     main = "HIE Participation by Bed Size",
     xlab = "(Log) Bed Size",
     ylab = "HIE Participation",
     col = "mistyrose2") # just call size rather than increase by one bed; log horitzonal HIE on vertical 
```
Looking at HIE participation across ownership type, all boxes have overlap with non-for profits participating the most and all others have similar lower values 

```{r, echo = FALSE}
# boxplot of hie participation by ownership type
boxplot(hie_part ~ data2$ownership,
        main = "HIE Participation by Ownership",
        xlab = "Ownership Type",
        ylab = "HIE Participation",
        names = c("Gov. non-federal", "Not-for-profit", "For-profit", "Gov. federal"),
        cex.axis = .6,
        col = "peachpuff")
```
Looking at HIE participation across EMR vendor, there is an obvious higher amount of participation for those who use Epic (more on why this makes sense later). Cerner has a slightly higher median level of HIE compared to other vendors or Meditech. 
```{r, echo = FALSE}
# boxplot of hie participation by primary ehr
boxplot(hie_part ~ data2$primary_emr,
        main = "HIE Participation by Primary EMR",
        xlab = "Primary EMR",
        ylab = "HIE Participation",
        names = c("Other", "Cerner", "Epic", "Meditech"),
        col = "skyblue1")
```
Now, looking at HIE participation by AI scores (with jitter as they are both slightly categorical although they are continuous). It does seem that those who participate in AI less also participate in HIE less and those who use AI more participate HIE more, so there could be a relationship there. 
```{r, echo = FALSE}
# scatterplot of hie particpation by ai score
plot(jitter(hie_part) ~ jitter(data2$ai_score), 
     main = "HIE Participation by AI Score",
     xlab = "AI Score",
     ylab = "HIE Participation",
     col = "lightsteelblue")
```

As a slight aside, I want to create a heat map of US States with darker states utilizing HIE more. HIE are historically regional, so this is very informative of what frameworks are still utilized. There is high participation in Michigan and Michigan and it seems to be darker on the East rather than the West. This could be reflective of more HIE structures or, because the average is being calculated, it could be because the Western states are larger and they have more hospitals whose data is being taken into account. 
```{r, echo = FALSE}
state_hie <- data2 %>%
  group_by(mstate) %>%
  summarise(avg_hie = mean(hie_participation, na.rm = TRUE)) %>%rename(state = mstate)

plot_usmap(data = state_hie, values = "avg_hie", regions = "states") +
  scale_fill_continuous(low = "white", high = "red", name = "HIE Participation") +
  labs(title = "Average HIE Participation by State") +
  theme(legend.position = "right")
```

# Analysis 

*Two Way T-Test*
I wanted to answer if non-profits and for-proift institutions participate in HIE the same on average?

Needed for a T-Test 
-make boxplot 
-% of values that are that category (potentially on top of barplot)
-boxplots for each group 
-summary statistics for each level of the variables. compare to raw mean and median
-t.test

There are 1065 non-profit organizations and 146 for-profit. 
```{r, include = FALSE}
# isolate only two groups
groups <- data2[data2$ownership == 2 | data2$ownership == 3, ]
table(groups$ownership) #1065 are non-profit; 146 are for profit
summary(groups$hie_participation)
by(groups$hie_participation, groups$ownership, summary)
```
Analyzing the boxplot, it seems not-for-profits participate more on average, but the confidence intervals do overlap. The overall mean and median (mean: 3.235 and median: 3) is closer to the mean and median for only not-for-profit (mean: 3.467, median: 3.000) than for-profits (mean: 1.541, median: 1.000). 

When we run the t test, we get a very small p value close to 0 meaning we can likely reject the null hypothesis that the mean is the same between the groups and conclude the mean HIE participation is not the same between not-for-profits and for-profits. From the confidence intervals, we can conclude that participation is higher in not-for-profits, which is what we expected from the boxplots. There are some high outliers in both groups. The spread both but especially the for-profit seem to be right skewed, although this would be better asessed with a histogram. 
```{r, echo = FALSE}
# raw summary statistics
summary(groups$hie_participation)

# summary statistics for each level
by(groups$hie_participation, groups$ownership, summary)

# make boxplot 
boxplot(groups$hie_participation ~ groups$ownership,
        main = "HIE Participation for \n Non-profit and For-profit",
        ylab = "HIE Participation",
        xlab = "Ownership Type",
        names = c("Not-for-profit", "For-profit"),
        col = "peachpuff")

# run the t test
t.test(groups$hie_participation ~ groups$ownership) # significant! with mean in group 2 being higher ie hospitals who are non-profits are more likely to participate in HIE
```


# Correlation
Are "bigger" hospitals (aka more beds) more likely to participate in HIE? As hospital size increases, does their participation in HIE?

Needed for a Correlation
-scatterplot (with correlation in the subhead) - ensure there is a linear relationship seen in scatterplot 
-add correlation value to subjeader

From the correlation test, we can see that there is a slight positive correlation between bed size and institution HIE participation, indicating bigger beds participate in HIE more, but this is not a particularly strong association. There does not seem to be any egregious outliers that are pulling the values, although we do know HIE participation is right skewed and may have some outliers.  
```{r, echo = FALSE}
# run correlation
cor.test(data2$hie_participation, log(data2$bdtot))

# correlation
cor <- round(cor(data2$hie_participation, log(data2$bdtot)), 2)

# plot with correlation in subheader and abline
plot(data2$hie_participation ~ log(data2$bdtot), 
     xlab = "(Log) Bed Size",
     ylab = "HIE Participation",
     col = "mistyrose3",
     pch = 19)
mtext(paste("Sample Correlation = ", cor), cex = .7, line =0)
mtext("HIE Participation by Bed Size", line = 1, font =  2)
```

I also wanted to looka at if hospitals who are using more AI, are participating in HIE more. Although these are both technically continuous variables, they are scored on a scale, so I added jitter to better appreciate frequency of answers. We see there is a moderate positive, correlation indicating institutions that utilize AI more perhaps utilize their HIE more. There are spots where answers cluster together, like those who use no AI also use HIE very little.   
```{r, echo = FALSE}
# run correlation
cor.test(data2$hie_participation, data2$ai_score)

# correlation
cor <- round(cor(data2$hie_participation, data2$ai_score), 2)

# scatterplot of hie particpation by ai score
plot(jitter(hie_part) ~ jitter(data2$ai_score), 
     xlab = "AI Score",
     ylab = "HIE Participation",
     col = "lightsteelblue")
mtext(paste("Sample Correlation = ", cor), cex = .7, line =0)
mtext("HIE Participation by AI Score", line = 1)
```

# Bootstrap
difference in HIE participation between ehr (not vs Epic - Epic vs not bc of cosmos)?

Needed for bootstrap of t-test:
-make barplot 
-% of values that are that category (potentially on top of barplot)
-boxplots for each group 
-summary statistics for each level of the variables. compare to raw mean and median
-t.test
-bootstrapping of mean (or metric measuring)
-histogram with line of where sample mean was and confidence intervals

What is the difference in HIE participation between EHR vendors, specifically those who use Epic versus those who do not. Epic has an added feature called Cosmos, which shares unidentified data with all institutions who use Epic and acts as a HIE. We would expect those who participate in Epic to have higher levels of HIE because of the availability of Epic to them. 

We see that even aggregating all non-Epic users together, the majority of the institutions still use Epic (47% are Epic users). There are 709 users not using Epic and 804 that are, so a good sample size for each. I want to bootstrap a t-test for this comparison, because looking at the boxplot, there are some high outliers in both groups and neither are normally distributed. 

Mean HIE participation for Epic users is higher (mean: 4.009) than non-Epic users (1.948) and closer to the overal average HIE participation scores (3.043). Running the parametric t-test, the pvalue is well small meaning we reject the null and conclude the means for each group are not the same. The We see from the bootstrap that the true difference is means is likely around 2.3 higher in Epic users, which is similar to what we saw in the parametric t-test. We also see that the boostrapped and parametric CIs are practically identical, indicating our original test was a good idea of the true distribution and not highly affected by outliers or skewed distribution. 
```{r, include = FALSE}
# create binary indicator of Epic or not 
data2$Epic <- ifelse(data2$primary_emr == 5, 1, 0)
range(data2$Epic) # check 0-1
```

```{r, echo = FALSE}
# barplot of each 
Epic <- table(data2$Epic)
pct <- paste0(round(Epic/ sum(Epic) * 100, 0), "%")
names(Epic) <- c("Not Epic", "Epic")
barplot(Epic[c(2, 1)],
        main = "Epic vs Not",
        ylim = c(0, 1000),
        names = c("Epic", "Not Epic"),
        col = c("red2", "red4"))
text(c(.75, 1.95), c(950, 820), pct, col = "seashell4", cex = 1)

# raw summary statistics
summary(data2$hie_participation)

# summary statistics for each level
by(data2$hie_participation, data2$Epic, summary)

# boxplot for each level
boxplot(data2$hie_participation ~ data2$Epic,
        names = c("Not Epic", "Epic"),
        col = c("red4", "red2"))
mtext("Mean with Epic: 4.01; Mean without Epic: 1.95", line = 0, cex = .7)
mtext("HIE Participation by Epic vs Not", line = 1, font = 2)
# first box is has Epic, second is does not

# t test
ttest <- t.test(data2$hie_participation ~ data2$Epic)

# bootstrap
n_samp <- 10000 # number of simulations want to run
diffMeans <- rep(NA, n_samp) # empty vector for sample mean differences

for(i in 1:n_samp){
  se <- sample(data2$hie_participation[data2$Epic == 1], sum(data2$Epic == 1), replace = T)
  so <- sample(data2$hie_participation[data2$Epic == 0], sum(data2$Epic == 0), replace = T)
  diffMeans[i] <- mean(se) - mean(so)
}

# add confidence intervals bootstratpped
ci <- quantile(diffMeans, c(0.025, 0.975))

# confidence intervals from parametric tes
cit <- abs(ttest$conf.int)
# because not Epic is listed first the CI was coming through negative. doing absolute puts it on the same playing field as the bootstrapped

# histogram of bootstrapped means
hist(diffMeans,
     main = "Bootstrapped Sample Means \n Difference Between Epic and Not",
     col = "red4",
     xlab = "HIE Participation",
     xlim = c(1.4, 2.4))
abline(v = cit, lwd = 3, col = "black")
abline(v = ci, lwd = 3, col = "green4", lty = 2)
legend("topleft", col = c("black", "green4"), legend = c("Original CI", "Boostrapped CI"), lty = c(1, 2), lwd = 3)
str(data2$Epic)
# we see the ci of the test and the boostrapped overlapping this indicated the the CLT was kicked in and we had enough data in our original sample to get a good idea of the potential true value
```


#Permutation Test
Needed for permutation test:
-get actual mean difference 
-shuffle labels and recalculate wanted metric
-histogram of fake sample with line of original test metric value 

For parametric test, I wanted to ensure that the conclusion was stable, so I shuffled the labels of being an Epic users or not and recalculated difference in group mean. We can see from the resulting histogram that it is very unlikely to see a difference in group mean this high under the null hypothesis. 
```{r, echo = FALSE}
# what is original means
actualdiff <- by(data2$hie_participation, data2$Epic, mean)
# original difference
actualdiff <- actualdiff[2] - actualdiff[1]
actualdiff

# do permutation
n <- 10000
diffvals <- rep(NA, n)

for (i in 1:n){
  fakelabel <- sample(data2$Epic) # shuffle labels 
  diffvals[i] <- mean(data2$hie_participation[fakelabel == 1]) - mean(data2$hie_participation[fakelabel == 0])
}

# create histogram with mark of sample value
hist(diffvals,
     main = "Permuted Sample Mean Differences \n in HIE Participation by Epic or Not",
     xlab = "HIE Participation",
     col = "red4",
     xlim = c(-.5, 2.5))
abline(v = actualdiff, lty = 2, lwd = 3)
text(actualdiff - .1, 800, paste("Actual Diff in Means =", round(actualdiff, 2)), srt = 90, cex = .7)
```


# Multiple Regression 
Does bed amount, ownership type, primary emr used, and AI usage predict HIE participation?

Needed for multiple regression - Module 15
-Sissyphys steps
1. identify predictors and response
2. check relationships with scatterplots (pairsJDRS() will give good matrix plot)
3. fit model
4. best steps? remove insignificant predictors
5.check model assumptions of LINE (L-linearity; I-independence; N- normality of errors; E-qual variance of errors) with myRes plots

I wanted to assess if you could predict HIE participation based on certain variables: bed total(in log), ownership type, primary EMR used, and AI score. I used JDRS 


Because my HIE participation score is skewed right, I checked the relationship between all predictors and the outcome and predictor using JDRS matrix plot for the raw HIE participation score and a log HIE participation score. Prior to taking the log of the participation score, I added a constant of 0.5 because there were some zeros in my data where people did not participate in HIE at all. Because this then made them less than 1, there were some negative values. To combat this, I added 1.5 as a constant instead. I checked the relationship for all 3 differnet HIE. Ultimately, the original HIE values did not have high evidence of heteroskedasticity or non-linearity, but it did have the non-normality. Neither of the log transformation changed the scatterplots of predictors vs outcomes a lot and neither introduced heteroskedasticity, non-linear relationships, or outlier, but it did fix the right skewedness. 

Because I did not like how the log transformation made the predictors vs outcome scatterplots look, I went ahead and fit 3 different models each with a different type of HIE participation score (raw, log with constant .5 and log with constant 1.5). Comparing the AIC, the log HIE models have a significantly lower AIC. Between the two, the AIC is identical, so I took a Box Cox transformation to decide which to go with. The model of log with a constant of 1.5 has a lambda very close to 1 while the model of log with a constant .5 has a lambda close to .5. Because of this, I decided to continue forward with the models using log constant of 1.5. 

I rechecked the pairs for this and there is no evidence of heteroskedasticity, large outliers, or a non-linear relationship between the predictors and outcome. There is also no evidence of any of the above between the predictors nor is there evidence of large amounts of high correlation suggesting multicollinearity. 

```{r, warnings = FALSE}
# isolate wanted variables into subdata frame
df <- data2[ , c("hie_participation", "bdtot", "ownership", "primary_emr", "ai_score")]
df$bdtot <- log(df$bdtot) # make log
names(df)[names(df) == "bdtot"] <- "logbdtot" # rename column for consistency

# make categorical variables factors
df$ownership <- factor(data2$ownership)
df$primary_emr <- factor(data2$primary_emr)

df <- df[ , c("hie_participation", "logbdtot", "ownership", "primary_emr", "ai_score")]

# check relationship with pairsJDRS()
pairsJDRS(df)

# create log variable
df$pos_hie_participation <- df$hie_participation + .5 
df$log_hie <- log(df$pos_hie_participation)
df <- df[c("log_hie", "logbdtot", "ownership", "primary_emr", "ai_score")]

# check relationships again 
pairsJDRS(df)

# try adding a higher constant before taking log
df$pos_hie_participation <- data2$hie_participation + 1.5 
df$log_hie <- log(df$pos_hie_participation)
df <- df[c("log_hie", "logbdtot", "ownership", "primary_emr", "ai_score")]

# check relationships again 
pairsJDRS(df)

```



```{r, include = FALSE}
# fit model (glm) with raw hie 
df$hie_participation <- data2$hie_participation
df <- df[c("hie_participation", "log_hie", "logbdtot", "ownership", "primary_emr", "ai_score")]
lm <- glm(df$hie_participation ~ df$primary_emr + df$ownership + df$ai_score + df$logbdtot)
summary(lm)
```

```{r, include = FALSE}
# fit positive 1.5 log model 
lm1 <- glm(df$log_hie ~ df$primary_emr + df$ownership + df$ai_score + df$logbd)
summary(lm1)
# remove bed total (p value .9)
boxCox(lm1)
```


```{r, include = FALSE}
# fit model with .5 log 
df$log5_hie <- df$hie_participation + .5
lm2 <- glm(df$log5_hie ~ df$primary_emr + df$ownership + df$ai_score + df$logbd)
summary(lm2)
boxCox(lm2)
```

```{r, echo = FALSE}
par(mfrow = c(1, 3))
boxCox(lm1)
boxCox(lm2)
```

Now, I need to fit the model and determine what predictors are significant. I attempted forward regressiond and fit all predictors to the model. I first removed log bed total as it had the highest p value of .9. This was surprising as I thought originally beginning the question this would be a very good predictor, but we saw in the correlation above that they were not actually that strongly related. 
After removing log bed count, primary emr 2 - cerner, primary emr 5 - epic, ownership 3 - pro-profi, and ai score were all significant. This means I leave all other predictors in the model. The overall model is significant with a pvalue close to 0 and an adjusted R2 of .3926. This indicates this model can explain almost 40% of the variability in the data. 
```{r, echo = FALSE}
# log constant 1.5 as model forward regression
lm1 <- lm(df$log_hie ~ df$primary_emr + df$ownership + df$ai_score)
summary(lm1)
Anova(lm1, type = 3)
```

Lastly, I check the residuals of the model. The errors are overall normally distributed although there is some departure at the top right of the tail. This could be due to the continuous predictors being practically categorical and no variables having negative values. I did check the residual plots of the other tested models, and this one has the closest to normal distribution of errors. Additionally, there is no evidence of heterskedasticity on the fits vs studentized residuals plot, but you can see hte harsh bottom line which is typical of variables that do not extend past 0. There are a few leverage points but none to be worried about.
# DO THE CALCULATION FOR COOKs
```{r, echo = FALSE}
# residuals
myResPlots(lm1, "HIE")
```

```{r, include = FALSE}
# residuals of other models - just to check 
myResPlots(lm2, "HIE")
myResPlots(lm, "HIE")
```

# ANCOVA
HIE score by size and emr provider and ai score; is there an effect of ai score depending on emr

Needed for ANCOVA - Module 20
-check myResplots for continuous variables
-fit linear regression lm(hie ~ ai + emr + ai*emr); get summary
-run Anova with type 3 errors
-could do the plot of fitting different lines for different emr providers 
-best subsets of model
-check residuals

Lastly, I did an ANCOVA to see if there is an interaction between emr provider and AI score and how they affect HIE score. 

```{r, echo = FALSE}
# myRes for continuous variables

# fit linear regression
ancova1 <- lm(df$log_hie ~ df$ai_score + df$primary_emr + df$ai_score*df$primary_emr) # log hie based on ai and emr

summary(ancova1)
Anova(ancova1, type = 3)

# make plot of the best model: HIE by AI and EMR
coefs <- coef(ancova1)
line_cols <- c("lightsteelblue", "darkolivegreen", "tomato", "navajowhite")
plot(df$log_hie ~ df$ai_score, 
     col = line_cols[df$primary_emr], 
     pch = 16, 
     cex = 0.5,
     ylab = "HIE participation",
     xlab = "AI Score",
     main = "HIE Participation Modeled by \n AI Score and Primary EMR")
legend("topleft", col = line_cols, legend = c("Other", "Cerner", "Epic", "Meditech"), pch =16, cex = .7, horiz = TRUE)
abline(a = coefs[1], b = coefs[2], col = "lightsteelblue", lwd = 2)
for (i in 3:5){
  abline(a = coefs[1] + coefs[i], b = coefs[2] + coefs[i+3], col  = line_cols[i-1], lwd = 2)
}

# my Resplots
myResPlots(ancova1, "HIE, AI, EMR")
```

# Conclusion and Summary 

