---
title: "Final Project - Data Exploration"
author: "Caroline Gheen"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Project Requirements
1) Data Cleaning – demonstrate use of find/replace, data cleaning, dealing with missing 
values, text character replacement, matching.  It’s ok if your data didn’t require much of 
this. 
2) Graphics – somewhere in your analysis, show appropriate use of at least ONE of each of 
the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be 
part of residuals discussion) histogram. 
3) Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for 
either a t-test or a correlation. 
4) Permutation Test – include at least one. 
5) Multiple Regression – use either backwards stepwise regression or some form of best 
subsets regression.  Should include residual plots.  A GLM with a mix of continuous and 
categorical predictors is fine here.  This includes a clear discussion of the interpretation 
of the results of your model. 
6) ONE OF THE FOLLOWING TECHNIQUES – ANOVA (with some form of multiple comparisons and preferably with a non-parametric analogue for comparison or Welch’s 
ANOVA – include discussion of interaction plots and interaction terms if two-way), 
ANCOVA, Binary Logistic Regression, Multinomial Logistic Regression. 

## Plan of Attack
*FROM 2024*
AHA ID Number
Variables: bed total (continuous) 
state(categorical)

ownership (categorical- *Government, nonfederal*:12 = State; 13 = County; 14 = City; 15 = City-county; 16 = Hospital district or authority. *Nongovernment, not-for-profit*: 21 = Church operated; 23 = Other. *Investor-owned, for-profit*: 31 = Individual; 32 = Partnership; 33 = Corporation. *Government, federal*: 40 = Department of Defense; 44 = Public Health Service other than 47; 45 = Veterans Affairs; 46 = Federal other than 41-45, 47-48; 47 = Public Health Service Indian Service; 48 = Department of Justice) 

primary inpatient EHR (categorical - 1 = Allscripts; 2 = Cerner/Oracle Health; 3 = eClinical works; 4 = Eclipsys ;5 = Epic; 6 = GE; 7 = McKesson; 8 = MED3000; 9 = Meditech; 10 = NextGen; 11 = Harris Healthcare/QuadraMed;; 12 = Sage; 13 = Siemens; 14 = Self-developed; 15 = Other (specify)
*Other described*: 16 = Would prefer not to disclose; 17 = CPSI/Evident; 18 = HMS; 19 = Healthland; 20 = Vitera/Greenway; 21 = Evident; 22 = MEDHOST; 24 = Prognosis; 25 = Athenahealth; 26 = MedWorx; 27 = Health Care System; 28 = Azalea Health/Prognosis; 29 = Altera Digital Health; 30 = We do not use an EMR/EHR system
);

?Do you collect SDOH
?How collect SDOH - HIE is an option

level of participation in HIE (categorical - 1 = HIE/HIO is operational in my area and we are participating and actively exchanging data in at least one HIE/HIO; 2 = HIE/HIO is operational in my area but we are not participating; 3 = HIE/HIO is not operational in my area; 4 = Do not know); 

which HIE acteively participate in (categorical)
a. CommonWell Health Alliance		Q11_A	CWHA
b. e-Health Exchange		Q11_B	HLTHXCH
c. Civitas (formerly SHIEC)/Patient Centered Data Home (PCDH)		Q11_C	SHIEC 
d. Carequality (network-to-network trust/governance framework		Q11_D	SPCAREQ
e. EHR vendor-based-network that enables record location and exchange within the EHR vendor's network of users (e.g., Epic's Care Everywhere)		Q11_E	VENNET
f. Other national health information exchange network		Q11_F	OTHNET
Other national health information exchange network description		Q11_F_OTH	OTHNET_DESC
g. Do not participate in any national health information exchange networks (either via vendor or directly)		Q11_G	NETNO
h. Do not know (national health information exchange network)		Q11_H	NETDNK
			
Key			
1 = Yes			
0 = No			

?How does hospital submit data to PH reporting for each activity (HIE option)
?For each type of reporting does hospital use automated, manual or mix (binaryish)
?What challenges has hospital experienced in past year when sending information electronically to PH? (categorical broken down by type of report)

?Do you use predictive models/ML (categorical - yes/no)
?For which do you applied ML or predictive (categorical)

*Would have to use non 2024 for longitudinal data

# Required
Question: What hopsital characteristics (state, bed size, type, EHR vendor, ML use, PHH challenges?) predict active HIE participation? 
In 2024, what hospital characteristics are associated with higher levels of national HIE network participation?

Scatterplot: participation 
Boxplot: participation in HIE by state; participation in state by type of ownerhsip
Quantile: multiple regression
Histogram: multiple regression
T-test: 
Correlation: bed total vs HIE participation? As hospitals are bigger are they more likely to participate in HIE; Sum types of ML doing, does ML use increase at bigger hospital - is there an interaction with those who participate in HIE?
Bootstrap of T-test or correlation:
Permutation:
Multiple Regression:
ANCOVA:
Interaction plot:


```{r}
data <- read.csv("AHA - Query.csv")
dim(data) # 25899, 28
names(data) 
head(data)
```

```{r}
sum(is.na(data$bdtot)) # 26 - bed total

sum(is.na(data$mcntrl)) # 2 - control/ownership

sum(is.na(data$piemro)) # 0 - all values present - primary inpatient EHR vendor

sum(is.na(data$nm1oth)) # 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name1 Other

sum(is.na(data$rhio_lvl1)) # 878 missing  Level of HIE or HIO participation

sum(is.na(data$hie_noex)) # - Do not exhange (HIE Vendor)
nrow(data) - sum(is.na(data$hie_noex)) # 356 values included

sum(is.na(data$hie_nodis)) # (hie_nodis) Prefer not to disclose (HIE Vendor
nrow(data) - sum(is.na(data$hie_nodis)) # 98 values included
# can we assume those who did not say prefer not to disclose or do not exchange have a HIE?
# maybe I include all answers and them manually bunch all together - do only for one year

sum(is.na(data$emoreg)) #  Regional, state, or local health information exchange organization (HIE/HIO). Not local, proprietary, enterprise network)
nrow(data) - sum(is.na(data$emoreg)) # 11687 values included

sum(is.na(data$nmref1)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 1
nrow(data) - sum(is.na(data$nmref1)) # 5620 values included 

sum(is.na(data$year1)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 1
nrow(data) - sum(is.na(data$year1)) # 5606 values included

sum(is.na(data$nmref2))#  Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 2
nrow(data) - sum(is.na(data$nmref2)) #1681 values included

sum(is.na(data$year2)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 2
nrow(data) - sum(is.na(data$year2)) # 2046 values included

sum(is.na(data$nm2oth))# 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name2 Other

sum(is.na(data$nmref3)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 3
nrow(data) - sum(is.na(data$nmref3)) # 832 values included

sum(is.na(data$year3)) #Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 3
nrow(data) - sum(is.na(data$year3)) #1160 values included 

sum(is.na(data$nm3oth)) # 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name3 Other

sum(is.na(data$ssrshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Syndromic surveillance reporting
nrow(data) - sum(is.na(data$ssrshie)) # 2060 values included 
sum(is.na(data$irrshie)) #  If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Immunization registry reporting
nrow(data) - sum(is.na(data$irrshie)) # 2037 values included

sum(is.na(data$ecrshie)) # f you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Electronic case reporting
nrow(data) - sum(is.na(data$ecrshie)) #2006 values included

sum(is.na(data$phrshie)) #  If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Public health registry reporting
nrow(data) - sum(is.na(data$phrshie)) # 2005 values included 
sum(is.na(data$cdrshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Clinical data registry reporting
nrow(data) - sum(is.na(data$cdrshie)) # 2003 values included 
sum(is.na(data$erlshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Electronic reportable laboratory result reporting
nrow(data) - sum(is.na(data$erlshie)) # 2043 values included

sum(is.na(data$hcushie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Hospital capacity and utilization of medical supplies
nrow(data) - sum(is.na(data$hcushie)) # 1970 values included

sum(is.na(data$hiechrge)) # HIE charges your hospital additionally to submit data for public health reporting activities.
nrow(data) - sum(is.na(data$hiechrge)) # 2643 values included 

```

