---
title: "Final Project Work"
author: "Caroline Gheen"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warnings = FALSE}
# load packages
library(dplyr) # for piping
library(car) # for recoding
```

Load Initial data - qureried from AHA IT Supplemental Survey 2024. This is an annual survey sent out to all hospitals that are member of the American Hospital Association and asseses their technology and information technology status. It is required and typically filled out by institutions' CIO. 
```{r}
# load initial data
data <- read.csv("AHA Data - Query 2.csv")
dim(data) # 2641 institutions responded, only loaded 48 variables; in total 27 questions are asked
names(data)
```

Change names to be more human readable and track question number to understand which go together
Many questions are the same question by binary indicators of which feature the institution does as part of that question. 
```{r}
# change names to be more human readable
names(data)[names(data) == "piemro"] <- "primary_emr_other"
names(data)[names(data) == "mcntrl"] <- "ownership"
names(data)[names(data) == "piemr"] <- "primary_emr"
names(data)[names(data) == "rhio_lvl1"] <- "q10_HIE_partic"
names(data)[names(data) == "othnet_desc"] <- "q11_other_nat_desc"
names(data)[names(data) == "cwha"] <- "q11_commonwell"
names(data)[names(data) == "hlthxch"] <- "q11_ehealth"
names(data)[names(data) == "netdnk"] <- "q11_dnk_othernat"
names(data)[names(data) == "netno"] <- "q11_no_nat"
names(data)[names(data) == "othnet"] <- "q11_other_nat"
names(data)[names(data) == "shiec"] <- "q11_civitas"
names(data)[names(data) == "spcareq"] <- "q11_carequality"
names(data)[names(data) == "vennet"] <- "q11_ehrbased"
names(data)[names(data) == "ssrstrdt"] <- "q15_syndromic"
names(data)[names(data) == "irrstrdt"] <- "q15_immune"
names(data)[names(data) == "ecrstrdt"] <- "q15_electronic"
names(data)[names(data) == "phrstrdt"] <- "q15_phreg"
names(data)[names(data) == "cdrstrdt"] <- "q15_clinicreg"
names(data)[names(data) == "erlstrdt"] <- "q15_elab"
names(data)[names(data) == "hcustrdt"] <- "q15_hospcap"
names(data)[names(data) == "ssrhie"] <- "q14_syndromic_hie"
names(data)[names(data) == "iirhie"] <- "q14_immune_hie"
names(data)[names(data) == "ecrhie"] <- "q14_electronic_hie"
names(data)[names(data) == "phrhie"] <- "q14_phreg_hie"
names(data)[names(data) == "cdrhie"] <- "q14_clinicreg_hie"
names(data)[names(data) == "erlhie"] <- "q14_elab_hie"
names(data)[names(data) == "hcuhie"] <- "q14_hospcap_hie"
names(data)[names(data) == "amichie"] <- "q14_amr_hie"
names(data)[names(data) == "aitraj"] <- "q19_ai_traj"
names(data)[names(data) == "airfol"] <- "q19_ai_risk_fu"
names(data)[names(data) == "aimhea"] <- "q19_ai_monitor"
names(data)[names(data) == "airect"] <- "q19_ai_rec"
names(data)[names(data) == "aibill"] <- "q19_ai_bill"
names(data)[names(data) == "aische"] <- "q19_ai_schedule"
names(data)[names(data) == "aipoth"] <- "q19_ai_process"
names(data)[names(data) == "aipoth_desc"] <- "q19_ai_process_desc"
names(data)[names(data) == "aicloth"] <- "q19_ai_clinic"
names(data)[names(data) == "aicloth_desc"] <- "q19_ai_clinic_desc"
names(data)[names(data) == "ainota"] <- "q19_ai_none"
names(data)[names(data) == "aidnk"] <- "q19_ai_dnk"
names(data)[names(data) == "aurstrdt"] <- "q15_amr"
names(data)[names(data) == "predml"] <- "q18_ml"
names(data)[names(data) == "predoth"] <- "q18_otherml"
data$othpad_desc <- NULL
names(data)
```

Explore data within each column for missingness
```{r}
# make df so easily see for each column how many missing, complete and % complete 
na_summary <- data.frame(
  n_data = colSums(!is.na(data)),
  n_na = colSums(is.na(data)),
  pct_comp = round(colSums(!is.na(data)) / nrow(data) * 100, 1)
)
na_summary
# there is lots of really good data - not a lot of missingness for rows
# I wonder if the missingness to the rows are coming from the same organizations ie some were really thorough in filling out the survey and some were not

# look at rows- if same organizations are NA for all, will drop them
row_na_summary <- data.frame(
  id = data$id,
  n_data = rowSums(!is.na(data)),
  n_na = rowSums(is.na(data)),
  pct_comp = round(rowSums(!is.na(data)) / ncol(data) * 100, 1)
)
row_na_summary

# just looking a the first few row entries, a lot of organizations have 100% complete, some have 90-100%, even less have 60-90%, and some have around 20%. I definitely want to drop the ones that only have 20% of data complete, but how harsh can I be?
# what is my nrow difference (ie how many institutions I retain) if I make a harsh boundary of only keep those who completed 90%+ vs those who completed 60%+. Or even if I only kept those who answered all questions
nrow(row_na_summary[row_na_summary$pct_comp > 61, ]) # 2485
nrow(row_na_summary[row_na_summary$pct_comp > 90, ]) # 2098
nrow(row_na_summary[row_na_summary$pct_comp == 100.0, ]) # 1513

# select selects that are full as best
best_rows <-(row_na_summary[row_na_summary$pct_comp == 100.0, ])

# keep only data that has all rows full as data2
data2 <- data %>% filter(id %in% best_rows$id) 

# sanity check that there are no NA - there are empty values but these are meaningful (ie. no primary other emr means they don't use other primary emr)
sum(is.na(data2)) # sanity check - 0
```
Only keeping institutions that answered all my questions of interest I still retain 1513 institutions. Because this is still a high number, I am going to institute this harsh boundary and only analyze those who completed all questions. This means moving forward I will use data4 and retain data as the original, ground truth dataset in case of any errors. 

Now to explore data. 
```{r}
str(data2)
```

Some questions have No keyed as 0 and some have no keyed as 2...
Would like to recode so it is consistant throughout entire data and I can make cumulative variables. I will recode everything so that 0=No, 1 = Yes...
This only matters for Q11 (already 0=No), Q14 (already 0=No), Q18 (currently 1=yes, 2=no, 3=do not know), Q19 (already 0=No), 
These need to be recoded. code as 0=no, 1=yes, 2=do not know. that means current 2 becomes 0, 1 stays 1, current 3 becomes 2
"q18_ml"              "q18_otherml" 
```{r}
# recode question 14
data2$q18_ml <- recode(data2$q18_ml, "2 = 0; 3 = 2" )

data2$q18_otherml <- recode(data2$q18_otherml, "2 = 0; 3 = 2" )

unique(data2$q18_ml)
```


Group together similar variables or create others (ie Epic, Cerner, Other)


First, let's explore the data a little more
```{r}
str(data2)
# structure seems good for all. my coding made them numeric instead of integer - this is probably fine. will likely need to make these factor to do categorical things. can come back and do this once I decide
```

Lets look at if all states are represented
```{r}
length(unique(data2$mstate)) # 52 states - PR and MP in addition to all 50 states

# I have no idea what MP is...
data2[data2$mstate == "MP",]
# googling the name of the health center and MP I discover it is in Mariana Islands for Navy members stationed there

# okay so we know all 50 state plus some are represented - how many hospitals are in each
table(data2$mstate)
# make this into a df so it is plottable
state_counts <- as.data.frame(table(data$mstate))
colnames(state_counts) <- c("state", "n")

barplot(state_counts$n)
hist(state_counts$n) # this is right skewed as is expected
# it would be cool to come back and do some sort of frequency plot that makes the state name bigger the more hospitals it has
```

Let's look at the type of primary EHRs that are used 
```{r}
table(data2$primary_emr)
# lots at Cerner/Oracle(2) and Epic (5) and Meditech(9) as expected. Also lots ie 50+ at Other(91 - these are in the other variable), CPSI/Evident(17).

as.data.frame(table(data2$primary_emr_other))
# Wellsky has a lot of responses (32)
# MyAvatar that is also NetSmart has lots of responses but they aren't going together based on different spelling and capitalization - I don't think it will add together high enough to become it's onw category so i don't think i'll spend the time to standardize

length(unique(data2$primary_emr_other)) # only 37 specified the other system they used but 91 said they use another system...
```
Cerner/Oracle, Epic, and Meditech have the most users as expected. I am going to do these three and all others will be an other, except 30 which is don't have an EMR and should stay as it's own category.Acutally, nobody answered 30 meaning everyboyd has an EMR
```{r}
# condense primary emr answers
data2$primary_emr[!data2$primary_emr %in% c(2, 5, 9, 30)] <- 0

# check only 2, 5, 9, and 30 are options for emr
unique(data2$primary_emr)

# new table with condensed options
table(data2$primary_emr)
```

Regroup ownership type variable
```{r}
# look at ownership as is
table(data2$ownership)
table(data$ownership)
# 12-16 is all types of government-nonfederal; 21 and 23 are non-government/not for profits; 31-33 are for profits; 40-48 are federal goverment types 
# not sure why an ownership option is the department of justice but none of those appear in this dataset - maybe if they go bankrupt 

# I am going to combine the individual breakdowns into the larger general groups given in the dataset
# government non-federal = 1; nongov, not for profit = 2; for profit = 3; federal = 4
data2$ownership[data2$ownership %in% c(12, 13, 14, 15, 16)] <- 1

data2$ownership[data2$ownership %in% c(21, 23)] <- 2

data2$ownership[data2$ownership %in% c(31, 32, 33)] <- 3

data2$ownership[data2$ownership %in% c(40, 44, 45, 46, 47, 48)] <- 4

# check current unique values 
unique(data2$ownership)
```
All types of nonfederal government hospital ownership groups are represented as well as for profit and not for profit. Not for profits are the largest category, altoghout other is larger than church, church alone is larger than some toehr groups. Federal is smaller than expected, there is only 1 owned by the DOD (military hospitals) and 9 VAs... this is not representative of the actual hospital ownership breakdown so this must be due to only keeping entries that answered all our questions. *Would be interesting to compare to owernship type in all of the dataset.*


I'm going to explore the rest of the description/free answer
 q11_other_HIE
 q19_ai_process_desc
 q19_ai_clinic_desc
```{r}
# 3 free text variables
as.data.frame(table(data2$q11_other_nat_desc)

as.data.frame(table(data2$q19_ai_process_desc))

as.data.frame(table(data2$q19_ai_clinic_desc))
```

All the other answers include the occurances of answerees leaving it empty. We need to drop these so they don't affect the frequency. We can replace all instances of no answer ("") with NA. 
```{r}
data2$q11_other_nat_desc[data2$q11_other_nat_desc == ""] <- NA
data2$q19_ai_clinic_desc[data2$q19_ai_clinic_desc == ""] <- NA
data2$q19_ai_process_desc[data2$q19_ai_process_desc == ""] <- NA

# recheck unique
unique(data2$q11_other_nat_desc)
unique(data2$q19_ai_clinic_desc)
unique(data2$q19_ai_process_desc)
# all "" changed to NA successfully
```


Explore Other HIE answers 
See if any of the occurrences in the free text are actually options
```{r}
# other HIE
as.data.frame(table(sort(data2$q11_other_nat_desc)))
# 1 for Commonwell (CommonWell through Healthix) - add to CommonWell; 6 for Epic - add to Epic; 3 for TEFCA - remove; TEFCA is an agreement not a framework; 1 eCQM - remove; eCQM is a CMS quality measure; 1 NIH - NIH does not have an HIE

# not sure if i want to have these i'm removing be added to do not know, do not use, or become NA
```
Lots of the other national exchanges are state exchanges that still exist - CA, CO, NJ, CT/MA, KY, IN KA, OH, OK; or other federal exchages lots with the VA; Surescripts and CyncHealth, Direct Trust also large chunk but not enough to challenge other groups to be their own 

Explore AI clinical other uses 
```{r}
as.data.frame(table(sort(data2$q19_ai_clinic_desc)))
```

Explore AI operational process other uses 
```{r}
as.data.frame(table(sort(data2$q19_ai_process_desc)))
```
Other clinical uses include a lot of predictive and LOS prediction that I think can be assigned out to a or f. Operational includes lots of supply chain which makes sense but also coding and revenue cycles that I think can be assigned out to billing. There doesn't seem to be a good difference of when to report something as other clinical process or other operational process (ie. clinical workflow and citation creation appear in both). Once I reassign out the ones I think I can, I am going to combine the variables to be one other variable for q19. 


Create subvariable for each question and frequency of answers
(can't do column sum on ones with 2/3 etc, but can create cleaner tables)
```{r}
# quesiton 11 - includes all except other frequency
q11df <- data2[, c("q11_carequality", "q11_commonwell", "q11_ehealth", "q11_dnk_othernat", "q11_no_nat", "q11_other_nat", "q11_civitas", "q11_ehrbased")]
q11_freq <- as.data.frame(colSums(q11df))
names(q11_freq) <- c("n")
# could expand these out to be a percentage - add another column (use dplyr) of n/total hospitals in dataset to create % of hospitals who are in that HIE type
# not all hospitals that said they use another HIE provided the name (~10, would be size of how many there are in that single variable(not whole column of the df bc that could be people who chose another column and thus left the please specify other column blank))

# question 19 
q19df <- data2[, c("q19_ai_traj","q19_ai_risk_fu", "q19_ai_monitor", "q19_ai_rec", "q19_ai_bill", "q19_ai_schedule", "q19_ai_process", "q19_ai_clinic", "q19_ai_none", "q19_ai_dnk")]
q19_freq <- as.data.frame(colSums(q19df))
names(q19_freq) <- c("n")
# could expand these out to be a percentage - add another column (use dplyr) of n/total hospitals in dataset to create % of hospitals who use that tool 

# question 14

q14df <- data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie")]
q14_freq <- as.data.frame(colSums(q14df))
names(q14_freq) <- c("n")

names(data2)

# this is all the variables that have binary
```

Need to create HIE participant cumulative variable. To score hospitals on their HIE participation, they will get +1 for each ph reporting they utilize HIE for, captured in question 14. Because these are already scored as 0 for no and 1 for yes, we can row sum these variables. Then, they will get +1 if they participate in an HIE. This would mean they answered 1 to question 10. Because no/other answers are the higher options for this question, we cannot just sum the answers. 
```{r}
data2$hie_participation <- rowSums(data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie")])

data2$hie_participation <- ifelse(data2$q10_HIE_partic == 1, data2$hie_participation + 1, data2$hie_participation)
```

Explore HIE participation variable
```{r}
# create own variable to cut down on typing
hie_part <- data2$hie_participation
range(hie_part) # 0-9 this is good
table(hie_part)

# create boxplot of hie participation by state
boxplot(hie_part ~ data2$mstate)

# scatterplot of hie participation by bed total
# most likely will want to take log transformation of bed total
hist(data2$bdtot)
hist(log(data2$bdtot))
plot(hie_part ~ log(data2$bdtot))

# scatterplot of hie participation by ownership type
hist(data2$ownership)
boxplot(hie_part ~ data2$ownership)

# scatterplot of hie participation by primary ehr
hist(data2$primary_emr)
boxplot(hie_part ~ data2$primary_emr)
```

