---
title: "Final Project Work"
author: "Caroline Gheen"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### ALL CODE
# Load Packages
```{r, warning = FALSE}
# load packages
library(dplyr) # for piping
library(car) # for recoding
library(usmap) # for heatmap
library(ggplot2) # for graphing
source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt") # source for pairsJDRS and MyResPlots
```

# Load Original Data
Load Initial data - qureried from AHA IT Supplemental Survey 2024. This is an annual survey sent out to all hospitals that are member of the American Hospital Association and asseses their technology and information technology status. It is required and typically filled out by institutions' CIO. 
```{r, include = FALSE}
# load initial data
data <- read.csv("AHA Data - Query 2.csv")
dim(data) # 2641 institutions responded, only loaded 48 variables; in total 27 questions are asked
names(data)
str(data)
```

# Change Column Names
Change names to be more human readable and track question number to understand which go together
Many questions are the same question but binary indicators of which feature the institution does as part of that question, so I included that in the question. I did export some questions that were not ultimately used in a final analyses, but they were retained nonetheless for future use if wanted. 
```{r, include = FALSE}
# change names to be more human readable
names(data)[names(data) == "piemro"] <- "primary_emr_other"
names(data)[names(data) == "mcntrl"] <- "ownership"
names(data)[names(data) == "piemr"] <- "primary_emr"
names(data)[names(data) == "rhio_lvl1"] <- "q10_HIE_partic"
names(data)[names(data) == "othnet_desc"] <- "q11_other_nat_desc"
names(data)[names(data) == "cwha"] <- "q11_commonwell"
names(data)[names(data) == "hlthxch"] <- "q11_ehealth"
names(data)[names(data) == "netdnk"] <- "q11_dnk_othernat"
names(data)[names(data) == "netno"] <- "q11_no_nat"
names(data)[names(data) == "othnet"] <- "q11_other_nat"
names(data)[names(data) == "shiec"] <- "q11_civitas"
names(data)[names(data) == "spcareq"] <- "q11_carequality"
names(data)[names(data) == "vennet"] <- "q11_ehrbased"
names(data)[names(data) == "ssrstrdt"] <- "q15_syndromic"
names(data)[names(data) == "irrstrdt"] <- "q15_immune"
names(data)[names(data) == "ecrstrdt"] <- "q15_electronic"
names(data)[names(data) == "phrstrdt"] <- "q15_phreg"
names(data)[names(data) == "cdrstrdt"] <- "q15_clinicreg"
names(data)[names(data) == "erlstrdt"] <- "q15_elab"
names(data)[names(data) == "hcustrdt"] <- "q15_hospcap"
names(data)[names(data) == "ssrhie"] <- "q14_syndromic_hie"
names(data)[names(data) == "iirhie"] <- "q14_immune_hie"
names(data)[names(data) == "ecrhie"] <- "q14_electronic_hie"
names(data)[names(data) == "phrhie"] <- "q14_phreg_hie"
names(data)[names(data) == "cdrhie"] <- "q14_clinicreg_hie"
names(data)[names(data) == "erlhie"] <- "q14_elab_hie"
names(data)[names(data) == "hcuhie"] <- "q14_hospcap_hie"
names(data)[names(data) == "amichie"] <- "q14_amr_hie"
names(data)[names(data) == "aitraj"] <- "q19_ai_traj"
names(data)[names(data) == "airfol"] <- "q19_ai_risk_fu"
names(data)[names(data) == "aimhea"] <- "q19_ai_monitor"
names(data)[names(data) == "airect"] <- "q19_ai_rec"
names(data)[names(data) == "aibill"] <- "q19_ai_bill"
names(data)[names(data) == "aische"] <- "q19_ai_schedule"
names(data)[names(data) == "aipoth"] <- "q19_ai_process"
names(data)[names(data) == "aipoth_desc"] <- "q19_ai_process_desc"
names(data)[names(data) == "aicloth"] <- "q19_ai_clinic"
names(data)[names(data) == "aicloth_desc"] <- "q19_ai_clinic_desc"
names(data)[names(data) == "ainota"] <- "q19_ai_none"
names(data)[names(data) == "aidnk"] <- "q19_ai_dnk"
names(data)[names(data) == "aurstrdt"] <- "q15_amr"
names(data)[names(data) == "predml"] <- "q18_ml"
names(data)[names(data) == "predoth"] <- "q18_otherml"
data$othpad_desc <- NULL
```

```{r}
names(data)
```

# Explore Missingness
Explore data within each column for missingness
```{r, include = FALSE}
# make df so easily see for each column how many missing, complete and % complete 
na_summary <- data.frame(
  n_data = colSums(!is.na(data)),
  n_na = colSums(is.na(data)),
  pct_comp = round(colSums(!is.na(data)) / nrow(data) * 100, 1)
)

# there is lots of really good data - not a lot of missingness for columns
# I wonder if the missingness to the columns  are coming from the same organizations ie some were really thorough in filling out the survey and some were not

# look at rows- if same organizations are NA for all, will potentially drop them
row_na_summary <- data.frame(
  id = data$id,
  n_data = rowSums(!is.na(data)),
  n_na = rowSums(is.na(data)),
  pct_comp = round(rowSums(!is.na(data)) / ncol(data) * 100, 1)
)
```

# COM BACK TO THIS TRY AND GET THEM IN ORDER FOR HEAD AND TAIL
```{r}
## include table
names(na_summary) <- c("Complete Data", "Missing Data", "Percent Complete")
names(row_na_summary) <- c("ID", "Complete Data", "Missing Data", "Percent Complete")

na_summary <- na_summary %>% arrange('Percent Complete')
head(na_summary)
tail(na_summary)

row_na_summary <- row_na_summary %>% arrange('Percent Complete')
head(row_na_summary)
tail(row_na_summary)
```

```{r}
## include head and tail
head(na_summary)
tail(na_summary)

head(row_na_summary)
```

```{r, include = FALSE}
# just looking a the first few row entries, a lot of organizations have 100% complete, some have 90-100%, even less have 60-90%, and some have around 20%. I definitely want to drop the ones that only have 20% of data complete, but how harsh can I be while still retaining a meaningful number of entries?
# what is my nrow difference (ie how many institutions I retain) if I make a harsh boundary of only keep those who completed 90%+ vs those who completed 60%+. Or even if I only kept those who answered all questions
nrow(row_na_summary[row_na_summary$pct_comp > 61, ]) # 2485
nrow(row_na_summary[row_na_summary$pct_comp > 90, ]) # 2098
nrow(row_na_summary[row_na_summary$pct_comp == 100.0, ]) # 1513

# select selects that are full as best
best_rows <-(row_na_summary[row_na_summary$pct_comp == 100.0, ])

# keep only data that has all rows full as data2
data2 <- data %>% filter(id %in% best_rows$id) 

# sanity check that there are no NA - there are empty values but these are meaningful (ie. no primary other emr means they don't use other primary emr)
sum(is.na(data2)) # sanity check - 0
```

```{r}
# we dropped 1128 institutions off of missingness. This means we  can do full comparisons (don't have to use pairwise comparisons, for example)
nrow(data) - nrow(data2)
```
Only keeping institutions that answered all my questions of interest I still retain 1513 institutions. Because this is still a high number, I am going to institute this harsh boundary and only analyze those who completed all questions. This means moving forward I will use data4 and retain data as the original, ground truth dataset in case of any errors. 

Now to explore data. 

# Recode all No Answers to be 0
Some questions have No keyed as 0 and some have no keyed as 2...
Would like to recode so it is consistant throughout entire data and I can make cumulative variables. I will recode everything so that 0=No, 1 = Yes...
This only matters for Q11 (already 0=No), Q14 (already 0=No), Q18 (currently 1=yes, 2=no, 3=do not know), Q19 (already 0=No), 
These need to be recoded. code as 0=no, 1=yes, 2=do not know. that means current 2 becomes 0, 1 stays 1, current 3 becomes 2
"q18_ml"              "q18_otherml" 
```{r, include = FALSE}
# recode question 14
data2$q18_ml <- recode(data2$q18_ml, "2 = 0; 3 = 2" )

data2$q18_otherml <- recode(data2$q18_otherml, "2 = 0; 3 = 2" )

unique(data2$q18_ml)
```


First, let's explore the data a little more
```{r, include = FALSE}
str(data2)
# structure seems good for all. my coding made them numeric instead of integer - this is probably fine. will likely need to make these factor to do categorical things. can come back and do this once I decide
```

# State Exploration
Lets look at if all states are represented
```{r}
length(unique(data2$mstate)) # 52 states - PR and MP in addition to all 50 states

# I have no idea what MP is...
data2[data2$mstate == "MP",]
# googling the name of the health center and MP I discover it is in Mariana Islands for Navy members stationed there
```

```{r, include = FALSE}
# okay so we know all 50 state plus some are represented - how many hospitals are in each
table(data2$mstate)
# make this into a df so it is plottable
state_counts <- as.data.frame(table(data$mstate))
colnames(state_counts) <- c("state", "n")

state_counts %>% arrange(desc(n))
# Texas has 266 hospitals, but the mean of hospitals per state is 49... the minimum is 1 hospital in a state (Mariana Islands)
summary(state_counts$n)

barplot(state_counts$n)
boxplot(state_counts$n)
hist(state_counts$n) # this is right skewed as is expected
# it would be cool to come back and do some sort of frequency plot that makes the state name bigger the more hospitals it has
```


```{r}
## include about states 
summary(state_counts$n)

hist(state_counts$n,
     main = "Histogram of Hospitals per State",
     xlab = "Amount of Hospitals per State",
     ylim = c(0, 35),
     col = "tomato2")
```


# Condense EHR variables 
Let's look at the type of primary EHRs that are used 
```{r, include = FALSE}
table(data2$primary_emr)
# lots at Cerner/Oracle(2) and Epic (5) and Meditech(9) as expected. Also lots ie 50+ at Other(91 - these are in the other variable), CPSI/Evident(17).

as.data.frame(table(data2$primary_emr_other))
# Wellsky has a lot of responses (32)
# MyAvatar that is also NetSmart has lots of responses but they aren't going together based on different spelling and capitalization - I don't think it will add together high enough to become it's onw category so i don't think i'll spend the time to standardize

length(unique(data2$primary_emr_other)) # only 37 specified the other system they used but 91 said they use another system...
```
Cerner/Oracle, Epic, and Meditech have the most users as expected. I am going to do these three and all others will be an other, except 30 which is don't have an EMR and should stay as it's own category. Acutally, nobody answered 30 meaning everybody has an EMR (yay all can receive medicare funding)

```{r}
## include 
hist(data2$primary_emr, 
     main = "What Primary EMRs Are Used",
     xlab = "Coding of Primary EMR",
     ylim = c(0, 1000),
     col = "skyblue1")
```


```{r, include = FALSE}
# condense primary emr answers
data2$primary_emr[!data2$primary_emr %in% c(2, 5, 9, 30)] <- 0

# check only 2, 5, 9, and 30 are options for emr
unique(data2$primary_emr)

# new table with condensed options
table(data2$primary_emr)
```

```{r}
## include 
primary_emr <- table(data2$primary_emr)
tabpct <- paste0(round(primary_emr / sum(primary_emr) * 100, 0), "%")
names(primary_emr) <- c("Other", "Cerner", "Epic", "Meditech")

barplot(primary_emr, 
        main = "What EMRs are Used (Condensed)",
        xlab = "EMRs",
        col = "skyblue3", 
        ylim = c(0, 1000))
text(c(.75, 1.97, 3.15, 4.35), c(350, 400, 920, 320), tabpct, cex = .8, col = "seashell4")
```

# Condense Ownership variables
Regroup ownership type variable
```{r, include = FALSE}
# look at ownership as is
table(data2$ownership)
table(data$ownership)
# 12-16 is all types of government-nonfederal; 21 and 23 are non-government/not for profits; 31-33 are for profits; 40-48 are federal goverment types 
# not sure why an ownership option is the department of justice but none of those appear in this dataset - maybe if they go bankrupt 

# I am going to combine the individual breakdowns into the larger general groups given in the dataset
# government non-federal = 1; nongov, not for profit = 2; for profit = 3; federal = 4
data2$ownership[data2$ownership %in% c(12, 13, 14, 15, 16)] <- 1

data2$ownership[data2$ownership %in% c(21, 23)] <- 2

data2$ownership[data2$ownership %in% c(31, 32, 33)] <- 3

data2$ownership[data2$ownership %in% c(40, 44, 45, 46, 47, 48)] <- 4

# check current unique values 
unique(data2$ownership)
```

```{r}
## include 
ownership <- table(data2$ownership)
tabpct <- paste0(round(ownership / sum(ownership) * 100, 0), "%")
names(ownership) <- c("Gov. non-federal", "Not-for-profit", "For-profit", "Gov. federal")

barplot(ownership,
        xlab= "Owners",
        main = "Ownership Types",
        ylim = c(0, 1300),
        cex.names = .6,
        col = "peachpuff3")
text(c(.75, 1.97, 3.15, 4.35), c(400, 1200, 250, 150), tabpct, cex = .8, col = "seashell4")
```
All types of nonfederal government hospital ownership groups are represented as well as for profit and not for profit. Not for profits are the largest category, altoghout other is larger than church, church alone is larger than some other groups. Federal is smaller than expected, there is only 1 owned by the DOD (military hospitals) and 9 VAs... this is not representative of the actual hospital ownership breakdown so this must be due to only keeping entries that answered all our questions.

# Bed Total
Bed Total. Used as proxy for size
```{r}
# scatterplot of hie participation by bed total
# most likely will want to take log transformation of bed total
hist(data2$bdtot, 
     main = "Total Number of Beds", 
     xlab = "Number of Beds",
     col = "mistyrose3")
hist(log(data2$bdtot),
     main = "Log Total Number of Beds",
     xlab = "Log Number of Beds",
     col = "mistyrose")
```

# Explore Free Text Answer Questions
I'm going to explore the rest of the description/free answer
 q11_other_HIE
 q19_ai_process_desc
 q19_ai_clinic_desc
```{r, include = FALSE}
# 3 free text variables
as.data.frame(table(data2$q11_other_nat_desc))

as.data.frame(table(data2$q19_ai_process_desc))

as.data.frame(table(data2$q19_ai_clinic_desc))
```

All the other answers include the occurances of answerees leaving it empty. We need to drop these so they don't affect the frequency. We can replace all instances of no answer ("") with NA. 
```{r, include = FALSE}
data2$q11_other_nat_desc[data2$q11_other_nat_desc == ""] <- NA
data2$q19_ai_clinic_desc[data2$q19_ai_clinic_desc == ""] <- NA
data2$q19_ai_process_desc[data2$q19_ai_process_desc == ""] <- NA

# recheck unique
#unique(data2$q11_other_nat_desc)
#unique(data2$q19_ai_clinic_desc)
#unique(data2$q19_ai_process_desc)
# all "" changed to NA successfully
```

Explore Other HIE answers 
See if any of the occurrences in the free text are actually options
```{r, include = FALSE}
# show full entries for any hospital that has an entry in other please specify 
data2[!is.na(data2$q11_other_nat_desc), ]

# other HIE
as.data.frame(table(data2$q11_other_nat_desc))
# 1 for Commonwell (CommonWell through Healthix) - they also marked commonwell so this will just become 0 and other descriptor will become NA
data2[which(data2$q11_other_nat_desc == "CommonWell through Healthix"), ] 
data2$q11_other_nat[which(data2$q11_other_nat_desc == "CommonWell through Healthix")] <- 0 

#6 for Epic - they also maked that they use an ehr based HIE, so this will just become 0 and other descriptor will become NA; all hospitals who said this are Carilion meaning this survey was likely filled out by the same person for mulitple institutions
data2[which(data2$q11_other_nat_desc == "Epic"), ]
data2$q11_other_nat[which(data2$q11_other_nat_desc == "Epic")] <- 0

# 3 for TEFCA - remove; TEFCA is an agreement not a framework; 1 eCQM - remove; eCQM is a CMS quality measure; 1 National Research Institute - NRI does not have an HIE (they share genomes but)- these will become I do not know
data2$q11_other_nat[which(data2$q11_other_nat_desc == "TEFCA")] <- 0
data2$q11_other_nat[which(data2$q11_other_nat_desc == "eCQM")] <- 0
data2$q11_other_nat[which(data2$q11_other_nat_desc == "National Research Institue")] <- 0

```
Lots of the other national exchanges are state exchanges that still exist - CA, CO, NJ, CT/MA, KY, IN KA, OH, OK; or other federal exchages lots with the VA; Surescripts and CyncHealth, Direct Trust also large chunk but not enough to challenge other groups to be their own 

Explore AI clinical other uses 
```{r, include = FALSE}
as.data.frame(table(sort(data2$q19_ai_clinic_desc)))
```

Explore AI operational process other uses 
```{r, include = FALSE}
as.data.frame(table(sort(data2$q19_ai_process_desc)))
```
Other clinical uses include a lot of predictive and LOS prediction that I think can be assigned out to a or f. Operational includes lots of supply chain which makes sense but also coding and revenue cycles that I think can be assigned out to billing. There doesn't seem to be a good difference of when to report something as other clinical process or other operational process (ie. clinical workflow and citation creation appear in both). Once I reassign out the ones I think I can, I am going to combine the variables to be one other variable for q19. 
I actually don't feel I have enough domain knowledge to be resoring these differnetly than how the CIO answered them. I do think there should perhaps be more guidance from the AHA in what sort of tasks go under what bucket (how long has this question been asked). But because the ultimate varibale being made from this is if you are using AI, one point will be given regardless of if they are using it for an 'other' reason or a directly assigned reason. 

# Create subvariables
Create subvariable for each question and frequency of answers for tables
(can't do column sum on ones with 2/3 etc, but can create cleaner tables)
```{r}
# quesiton 11 - includes all except other frequency
q11df <- data2[, c("q11_carequality", "q11_commonwell", "q11_ehealth", "q11_dnk_othernat", "q11_no_nat", "q11_other_nat", "q11_civitas", "q11_ehrbased")]
q11_freq <- as.data.frame(colSums(q11df))
names(q11_freq) <- c("n")
pct <- paste0(round((q11_freq$n/nrow(data2)*100), 0), "%")
q11_freq <- cbind(q11_freq, pct)
#  % of hospitals a part of that hie 

# question 19 
q19df <- data2[, c("q19_ai_traj","q19_ai_risk_fu", "q19_ai_monitor", "q19_ai_rec", "q19_ai_bill", "q19_ai_schedule", "q19_ai_process", "q19_ai_clinic", "q19_ai_none", "q19_ai_dnk")]
q19_freq <- as.data.frame(colSums(q19df))
names(q19_freq) <- c("n")
pct <- paste0(round((q19_freq$n/nrow(data2)*100), 0), "%")
q19_freq <- cbind(q19_freq, pct)
#  % of hospitals who use that tool 

# question 14
q14df <- data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie")]
q14_freq <- as.data.frame(colSums(q14df))
names(q14_freq) <- c("n")
pct <- paste0(round((q14_freq$n/nrow(data2)*100), 0), "%")
q14_freq <- cbind(q14_freq, pct)
#  % of hospitals who use hie for that ph reporting

# this is all the variables that have binary
```

% of hospitals that participate in specific HIE
```{r, include = FALSE}
#  % of hospitals that use that hie
rownames(q11_freq) <- c("CareQuality", "CommonWell", "e-Health Exchange", "Do Not Know", "No HIE", "Other", "Civitas", "EHR-based")
```

% of hospitals who use applied ML or predictive analysis for that task 
```{r, include = FALSE}
#  % of hospitals who use that tool 
rownames(q19_freq) <- c("Predicting Health Trajectors or Risk for Inpatients", "Identify High Risk Outpatients", "Monitor Health", "Recommend Treatments", "Simplify or Automate Billing", "Facilitate Scheduling", "Other Operation Process", "Other Clinical Uses", "No ML Use", "Do Not Know")
```

```{r, include = FALSE}
# % oh hospitals who use that hie for ph reporting
rownames(q14_freq) <- c("Syndromic Surveillance", "Immunization Registry", "Electronic Case Reporting", "Public Health Registry", "Clinical Data Registry Reporting", "Electronic Reportable Lab Reporting", "Hospital Capacity and Utilization of Medical Supplies", "Antimicrobial Use and Resistances")
```

```{r}
q11_freq
q14_freq
q19_freq
```


# HIE Participation Cumulative Variable
Need to create HIE participant cumulative variable. To score hospitals on their HIE participation, they will get +1 for each ph reporting they utilize HIE for, captured in question 14. Because these are already scored as 0 for no and 1 for yes, we can row sum these variables. Then, they will get +1 for each HIE they participate in. This would mean they have a 1 for any question 11, except do not know "q11_dnk_othernat" or "q11_no_nat"
```{r, include = FALSE}
data2$hie_participation <- rowSums(data2[, c("q14_syndromic_hie", "q14_immune_hie", "q14_electronic_hie", "q14_phreg_hie", "q14_clinicreg_hie", "q14_elab_hie", "q14_hospcap_hie", "q14_amr_hie", "q11_other_nat", "q11_commonwell", "q11_ehealth", "q11_civitas", "q11_carequality","q11_ehrbased")])
```

Explore HIE participation variable
```{r, include = FALSE}
# create own variable to cut down on typing
hie_part <- data2$hie_participation
range(hie_part) # 0-13; max would have been 14, so nobody uses everything
table(hie_part)
```

```{r}
## include 
summary(hie_part)

hist(hie_part, 
     main = "Distribution of HIE Participation",
     xlab = "HIE Participation",
     col = "navajowhite")

```
# AI Use Composite Score
```{r, include = FALSE}
data2$ai_score <- rowSums(data2[, c("q19_ai_traj", "q19_ai_risk_fu", "q19_ai_monitor", "q19_ai_rec", "q19_ai_bill", "q19_ai_schedule", "q19_ai_process", "q19_ai_clinic")])
range(data2$ai_score) # someone uses AI for all
table(data2$ai_score) # a lot of hospitals are still not using AI, but most are integrating multiple uses
barplot(table(data2$ai_score))

table(data2$q18_ml)
table(data2$q18_otherml)
#1090 should be using AI in some way according to question 19 (number of hospitals - those who answered 0 = 1090) but according to question 18, only 800-900 answered yes to those. which do i trust.
```
Should I make an indicator variable that is (uses AI/does not use AI)??? According to question 19, 1090 should be using AI in some way (number of hospitals - those who answered 0 = 1090) but according to question 18, only 800-900 answered yes to those. which do i trus - probably question 19 because the 

```{r}
# include 
hist(data2$ai_score, 
     main = "Frequency of AI Scores",
     xlab = "AI Scores",
     col = "lightsteelblue",
     ylim = c(0, 500))
```

Overal HIE Participation
85% of hospitals participate in an HIE
5.5% of hospitals have an active area HIE that they do not participate in 
4.4% of hospitals do not have an operational HIE in their area (this is probably bigger- you can't select this and another numbers - CHECK IF YOU CAN SELECT MULTPILE)
4.2% do not know their HIE participation
```{r, include = FALSE}
hie <- as.data.frame(table(data2$q10_HIE_partic))
hie$pct <- paste0(round(hie$Freq / sum(hie$Freq) * 100, 0), "%")
# the majority of hospitals do participate in an HIE
names(hie) <- c("HIE Status", "Frequency", "Percent")
rownames(hie) <- c("Area HIE is operational and do participate", "Area HIE is operational but do not participation", "Area HIE is not operational", "Do Not Know")

hie
```



# Comparing Variables to Each Other
HIE Participation by Bed Total, Ownership, Primary EMR, and AI Score
```{r}
plot(hie_part ~ log(data2$bdtot), 
     main = "HIE Participation by Bed Size",
     xlab = "(Log) Bed Size",
     ylab = "HIE Participation",
     col = "mistyrose2") # just call size rather than increase by one bed; log horitzonal HIE on vertical 

# boxplot of hie participation by ownership type
boxplot(hie_part ~ data2$ownership,
        main = "HIE Participation by Ownership",
        xlab = "Ownership Type",
        ylab = "HIE Participation",
        names = c("Gov. non-federal", "Not-for-profit", "For-profit", "Gov. federal"),
        cex.axis = .6,
        col = "peachpuff")

# boxplot of hie participation by primary ehr
boxplot(hie_part ~ data2$primary_emr,
        main = "HIE Participation by Primary EMR",
        xlab = "Primary EMR",
        ylab = "HIE Participation",
        names = c("Other", "Cerner", "Epic", "Meditech"),
        col = "skyblue1")

# scatterplot of hie particpation by ai score
plot(jitter(hie_part) ~ jitter(data2$ai_score), 
     main = "HIE Participation by AI Score",
     xlab = "AI Score",
     ylab = "HIE Participation",
     col = "lightsteelblue")
```

Create heat map of US States that darker states utilize an HIE more 
```{r}
state_hie <- data2 %>%
  group_by(mstate) %>%
  summarise(avg_hie = mean(hie_participation, na.rm = TRUE)) %>%rename(state = mstate)

plot_usmap(data = state_hie, values = "avg_hie", regions = "states") +
  scale_fill_continuous(low = "white", high = "red", name = "HIE Participation") +
  labs(title = "Average HIE Participation by State") +
  theme(legend.position = "right")
```

# Two Way T-Test
Do non-profits and for-proift institutions participate in HIE the same on average?

Needed for a T-Test 
-make boxplot 
-% of values that are that category (potentially on top of barplot)
-boxplots for each group 
-summary statistics for each level of the variables. compare to raw mean and median
-t.test

```{r, include = FALSE}
# isolate only two groups
groups <- data2[data2$ownership == 2 | data2$ownership == 3, ]
table(groups$ownership) #1065 are non-profit; 146 are for profit
```

```{r}
# summary statistics for each level
by(groups$hie_participation, groups$ownership, summary)

# make boxplot 
boxplot(groups$hie_participation ~ groups$ownership,
        main = "HIE Participation for \n Non-profit and For-profit",
        ylab = "HIE Participation",
        xlab = "Ownership Type",
        names = c("Not-for-profit", "For-profit"),
        col = "peachpuff")

# run the t test
t.test(groups$hie_participation ~ groups$ownership) # significant! with mean in group 2 being higher ie hospitals who are non-profits are more likely to participate in HIE
```


# Correlation
Are bigger hospitals more likely to participate in HIE? As hospital size increases, does their participation in HIE?

Needed for a Correlation
-scatterplot (with correlation in the subhead) - ensure there is a linear relationship seen in scatterplot 
-add correlation value to subjeader
```{r}
# run correlation
cor.test(data2$hie_participation, log(data2$bdtot))

# correlation
cor <- round(cor(data2$hie_participation, log(data2$bdtot)), 2)

# plot with correlation in subheader and abline
plot(hie_part ~ log(data2$bdtot), 
     xlab = "(Log) Bed Size",
     ylab = "HIE Participation",
     col = "mistyrose2")
mtext(paste("Sample Correlation = ", cor), cex = .7, line =0)
mtext("HIE Participation by Bed Size", line = 1)
```

Are hospitals who are using more AI, participating in HIE more?
```{r}
# run correlation
cor.test(data2$hie_participation, data2$ai_score)

# correlation
cor <- round(cor(data2$hie_participation, data2$ai_score), 2)

# scatterplot of hie particpation by ai score
plot(jitter(hie_part) ~ jitter(data2$ai_score), 
     xlab = "AI Score",
     ylab = "HIE Participation",
     col = "lightsteelblue")
mtext(paste("Sample Correlation = ", cor), cex = .7, line =0)
mtext("HIE Participation by AI Score", line = 1)
```

# Bootstrap
difference in HIE participation between ehr (not vs epic - epic vs not bc of cosmos)?

Needed for bootstrap of t-test:
-make barplot 
-% of values that are that category (potentially on top of barplot)
-boxplots for each group 
-summary statistics for each level of the variables. compare to raw mean and median
-t.test
-bootstrapping of mean (or metric measuring)
-histogram with line of where sample mean was and confidence intervals
```{r, include = FALSE}
# create binary indicator of epic or not 
data2$epic <- ifelse(data2$primary_emr == 5, 1, 0)
range(data2$epic) # check 0-1
```

```{r}
# barplot of each 
epic <- table(data2$epic)
pct <- paste0(round(epic/ sum(epic) * 100, 0), "%")
names(epic) <- c("Not Epic", "Epic")
barplot(epic[c(2, 1)],
        main = "Epic vs Not",
        ylim = c(0, 1000),
        col = c("red4", "red3"))
text(c(.75, 1.95), c(950, 820), pct, col = "seashell4", cex = 1)

# summary statistics for each level
by(data2$hie_participation, data2$epic, summary)

# boxplot for each level
boxplot(data2$hie_participation, data2$epic,
        names = c("Epic", "Not Epic"),
        col = c("red3", "red4"))
mtext("Mean with Epic: 4.01; Mean without Epic: 1.95", line = 0, cex = .7)
mtext("HIE Participation by Epic vs Not", line = 1, font = 2)
# first box is has epic, second is does not

# t test
t.test(data2$hie_participation, data2$epi)

# bootstrap
n_samp <- 10000 # number of simulations want to run
diffMeans <- rep(NA, n_samp) # empty vector for sample mean differences

for(i in 1:n_samp){
  se <- sample(data2$hie_participation[data2$epic == 1], sum(data2$epic == 1), replace = T)
  so <- sample(data2$hie_participation[data2$epic == 0], sum(data2$epic == 0), replace = T)
  diffMeans[i] <- mean(se) - mean(so)
}

# add confidence intervals from original and bootstratpped
ci <- quantile(diffMeans, c(0.025, 0.975))

# test Ci
cit <- abs(t.test(hie_participation ~ epic, data = data2)$conf.int)
# because not epic is listed first the CI was coming through negative. doing absolute puts it on the same playing field as the bootstrapped

# histogram of bootstrapped means
hist(diffMeans,
     main = "Bootstrapped Sample Means \n Difference Between Epic and Not",
     col = "red4",
     xlab = "HIE Participation")
abline(v = cit, lwd = 3, col = "black")
abline(v = ci, lwd = 3, col = "green4", lty = 2)
# we see the ci of the test and the boostrapped overlapping this indicated the the CLT was kicked in and we had enough data in our original sample to get a good idea of the potential true value
```


#Permutation Test
Needed for permutation test:
-get actual mean difference 
-shuffle labels and recalculate wanted metric
-histogram of fake sample with line of original test metric value 

There is a difference in epic vs not, so I can shuffle labels and see same result?
```{r}
# what is original means
actualdiff <- by(data2$hie_participation, data2$epic, mean)
# original difference
actualdiff <- actualdiff[2] - actualdiff[1]
actualdiff

# do permutation
n <- 10000
diffvals <- rep(NA, n)

for (i in 1:n){
  fakelabel <- sample(data2$epic) # shuffle labels 
  diffvals[i] <- mean(data2$hie_participation[fakelabel == 1]) - mean(data2$hie_participation[fakelabel == 0])
}

# create histogram with mark of sample value
hist(diffvals,
     main = "Permuted Sample Mean Differences \n in HIE Participation by Epic or Not",
     xlab = "HIE Participation",
     col = "red4",
     xlim = c(-.5, 2.5))
abline(v = actualdiff, lty = 2, lwd = 3)
text(actualdiff - .1, 800, paste("Actual Diff in Means =", round(actualdiff, 2)), srt = 90, cex = .7)
```


# Multiple Regression 
Does bed amount, ownership type, primary emr used, and AI usage predict HIE participation?

Needed for multiple regression - Module 15
-Sissyphys steps
1. identify predictors and response
2. check relationships with scatterplots (pairsJDRS() will give good matrix plot)
3. fit model
4. best steps? remove insignificant predictors
5.check model assumptions of LINE (L-linearity; I-independence; N- normality of errors; E-qual variance of errors) with myRes plots

```{r}
# isolate wanted variables into subdata frame
df <- data2[ , c("hie_participation", "bdtot", "ownership", "primary_emr", "ai_score")]
df$bdtot <- log(df$bdtot) # make log
names(df)[names(df) == "bdtot"] <- "logbdtot" # rename column for consistency

# make categorical variables factors
df$ownership <- factor(data2$ownership)
df$primary_emr <- factor(data2$primary_emr)

# check relationship with pairsJDRS()
pairsJDRS(df)
# this doesn't work bc i have somee categorical variables... i don't know if it's enough to have the graphs shown above or if i shoudl redo the scatterplot

# fit model (glm)
lm <- glm(df$hie_participation ~ df$primary_emr + df$ownership + df$ai_score)
summary(lm)

# best subsets
# biggest p value is logbdtotal (this feels surprising) - 0.67635
# reference for the categoricals is emr = 0 which is "other" and ownership 1 = gov, non-federal
#emr 2 (cerner - 0.00187) and 5 (epic - practically 0) are significant (not 9 - meditech - 0.59033)
# ownership 3 (for-profit - 0.02790) is significant but not 2 (not-for-profit - 0.33737) or 4 (federal - 0.52347)
# AI is very significant

Anova(lm)
# in ANOVA type 3, all are significant by 0.05; emr and AI are significant at 0.001

# check model assumptios with myRes()
myResPlots(lm, "HIE")
```


# ANCOVA
HIE score by size and emr provider and interaction; is there an effect of size depending on emr

Needed for ANCOVA - Module 20
-scattplot of continuous (accomplished above)
-check myResplots for continuous variables
-fit linear regression lm(hie ~ bed + emr + bed*emr); get summary
-run Anova with type 3 errors
-could do the plot of fitting different lines for different emr providers 
-best subsets of model
-check residuals

```{r}
# myRes for continuous variables

# fit linear regression
ancova <- lm(data2$hie_participation ~ log(data2$bdtot)*data2$primary_emr)
summary(lm)

# Anova
Anova(ancova, type = 3)

# my Resplots 
```

### Report
# Introduction
I utilized the 2024 American Hospital Association's Annual Survey IT supplement as my data source. This is an annual survey sent out to all hospitals that are member of the American Hospital Association and asseses their technology and information technology status. It is required and typically filled out by institutions' CIO.
I am studying Health Informatics at YSPH, so I am very interested in the information systems at hospitals and this is a dataset I have been wanting to explore for a while. I am interested in Health Information Exchange which is how hospitals exchange information to be better stewards of public health and better understand a patient so that their information can be passed to other doctors (look this up). I wanted to see if participation rates have been affected by electronic medical records expanding. 

# Data
I extracted a few questions from the survey: entry's AHA ID number; name; bed total (continuous range: ); state (categorical; 52 states - Puerto Rico and Marina Island Pacific are included); who owns them (categorical: government-nonfederal, not-for-profit, for-profit, governemnt-federal); what primary electronic health record is used (categorical including other free text entry); HIE 



Many questions are the same question but binary indicators of which feature the institution does as part of that question, so I included that in the question. 
1. AI Composite Score (continuous out of 8)
2. HIE Participation Score (continuous out of 14)
3. Number of Beds (continuous)
4. Ownership type (categorical)
5. Primary EMR (categorical)
6. State (categorical)

# Data Cleaning Process

# Descriptive Plots

# Analysis 

# Conclucaion and Summary 

