---
title: "Final Project - Data Exploration"
author: "Caroline Gheen"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Project Requirements
1) Data Cleaning – demonstrate use of find/replace, data cleaning, dealing with missing 
values, text character replacement, matching.  It’s ok if your data didn’t require much of 
this. 
2) Graphics – somewhere in your analysis, show appropriate use of at least ONE of each of 
the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be 
part of residuals discussion) histogram. 
3) Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for 
either a t-test or a correlation. 
4) Permutation Test – include at least one. 
5) Multiple Regression – use either backwards stepwise regression or some form of best 
subsets regression.  Should include residual plots.  A GLM with a mix of continuous and 
categorical predictors is fine here.  This includes a clear discussion of the interpretation 
of the results of your model. 
6) ONE OF THE FOLLOWING TECHNIQUES – ANOVA (with some form of multiple comparisons and preferably with a non-parametric analogue for comparison or Welch’s ANOVA – include discussion of interaction plots and interaction terms if two-way), ANCOVA, Binary Logistic Regression, Multinomial Logistic Regression. 

## Plan of Attack
*FROM 2024*
*ID*AHA ID Number
*MNAME*Name 
*BDTOT*Variables: bed total (continuous) 
*MSTATE*state(categorical)

*MCNTRL*ownership (categorical- *Government, nonfederal*:12 = State; 13 = County; 14 = City; 15 = City-county; 16 = Hospital district or authority. *Nongovernment, not-for-profit*: 21 = Church operated; 23 = Other. *Investor-owned, for-profit*: 31 = Individual; 32 = Partnership; 33 = Corporation. *Government, federal*: 40 = Department of Defense; 44 = Public Health Service other than 47; 45 = Veterans Affairs; 46 = Federal other than 41-45, 47-48; 47 = Public Health Service Indian Service; 48 = Department of Justice) 

*Q1 - PIEMR*primary inpatient EHR (categorical - 1 = Allscripts; 2 = Cerner/Oracle Health; 3 = eClinical works; 4 = Eclipsys ;5 = Epic; 6 = GE; 7 = McKesson; 8 = MED3000; 9 = Meditech; 10 = NextGen; 11 = Harris Healthcare/QuadraMed;; 12 = Sage; 13 = Siemens; 14 = Self-developed; 15 = Other (specify) 16 = Would prefer not to disclose; 17 = CPSI/Evident; 18 = HMS; 19 = Healthland; 20 = Vitera/Greenway; 21 = Evident; 22 = MEDHOST; 24 = Prognosis; 25 = Athenahealth; 26 = MedWorx; 27 = Health Care System; 28 = Azalea Health/Prognosis; 29 = Altera Digital Health; 30 = We do not use an EMR/EHR system
);
*Q2- PIERMO**Other described*:
Primary provider of inpatient EHR		Q1	PIEMR
Primary provider of inpatient EHR - other description		Q1_OTH	PIEMRO

*Q10 - RHIO_LVL1*level of participation in HIE (categorical - 1 = HIE/HIO is operational in my area and we are participating and actively exchanging data in at least one HIE/HIO; 2 = HIE/HIO is operational in my area but we are not participating; 3 = HIE/HIO is not operational in my area; 4 = Do not know); 
Level of HIE or HIO participation		Q10	RHIO_LVL1

*Q11 - CWHA; HLTHXCH; SHIEC; SPCAREQ; VENNET; OTHNET; OTHNET_DESC; NETNO; NETDNK*which HIE acteively participate in (categorical/binary - 1 = Yes; 0 = No)
a. CommonWell Health Alliance		Q11_A	CWHA
b. e-Health Exchange		Q11_B	HLTHXCH
c. Civitas (formerly SHIEC)/Patient Centered Data Home (PCDH)		Q11_C	SHIEC 
d. Carequality (network-to-network trust/governance framework		Q11_D	SPCAREQ
e. EHR vendor-based-network that enables record location and exchange within the EHR vendor's network of users (e.g., Epic's Care Everywhere)		Q11_E	VENNET
f. Other national health information exchange network		Q11_F	OTHNET
Other national health information exchange network description		Q11_F_OTH	OTHNET_DESC
g. Do not participate in any national health information exchange networks (either via vendor or directly)		Q11_G	NETNO
h. Do not know (national health information exchange network)		Q11_H	NETDNK
Key			
1 = Yes			
0 = No			

*Q14 only added those to do with HIE- SSREHR; SSRHIE; SSRPS; SSRFF; SSRNES; SSRDK; IIREHR; IIRHIE; IIRPS; IIRFF; IIRNES; IIRDK; ECREHR; ECRHIE; ECRPS; ECRFF; ECRNES; ECRDK; PHREHR; PHRHIE; PHRPS; PHRFF; PHRNES; PHRDK; CDREHR; CDRHIE; CDRPS; CDRFF; CDRNES; CDRDK; ERLEHR; ERLHIE; ERLPS; ERLFF; ERLNES; ERLDK; AMICEHR; AMICHIE; AMICPS; AMICFF; AMICNES; AMICDK; HCUEHR; HCUHIE; HCUPS; HCUFF; HCUNES; HCUDK* How does hospital submit data to PH reporting for each activity (binary - 1 = Yes; 0 = No)
a. Syndromic surveillance reporting - Directly from Electronic health record		Q14a_1	SSREHR
a. Syndromic surveillance reporting - Health information exchange		Q14a_2	# SSRHIE #
a. Syndromic surveillance reporting - Portal submission		Q14a_3	SSRPS
a. Syndromic surveillance reporting - Flat files 		Q14a_4	SSRFF
a. Syndromic surveillance reporting - Not electronically submitting		Q14a_5	SSRNES
a. Syndromic surveillance reporting - Do not know		Q14a_6	SSRDK
b. Immunization registry reporting - Directly from Electronic health record		Q14b_1	IIREHR
b. Immunization registry reporting - Health information exchange		Q14b_2	# IIRHIE #
b. Immunization registry reporting - Portal submission		Q14b_3	IIRPS
b. Immunization registry reporting - Flat files		Q14b_4	IIRFF
b. Immunization registry reporting  - Not electronically submitting		Q14b_5	IIRNES
b. Immunization registry reporting - Do not know		Q14b_6	IIRDK
c. Electronic case reporting - Directly from Electronic health record		Q14c_1	ECREHR
c. Electronic case reporting - Health information exchange		Q14c_2	# ECRHIE #
c. Electronic case reporting - Portal submission		Q14c_3	ECRPS
c. Electronic case reporting - Flat files		Q14c_4	ECRFF
c. Electronic case reporting - Not electronically submitting		Q14c_5	ECRNES
c. Electronic case reporting - Do not know		Q14c_6	ECRDK
d. Public health registry reporting - Directly from Electronic health record		Q14d_1	PHREHR
d. Public health registry reporting - Health information exchange		Q14d_2	# PHRHIE #
d. Public health registry reporting - Portal submission		Q14d_3	PHRPS
d. Public health registry reporting - Flat files		Q14d_4	PHRFF
d. Public health registry reporting - Not electronically submitting		Q14d_5	PHRNES
d. Public health registry reporting - Do not know		Q14d_6	PHRDK
e. Clinical data registry reporting - Directly from Electronic health record		Q14e_1	CDREHR
e. Clinical data registry reporting - Health information exchange		Q14e_2	# CDRHIE #
e. Clinical data registry reporting - Portal submission		Q14e_3	CDRPS
e. Clinical data registry reporting - Flat files		Q14e_4	CDRFF
e. Clinical data registry reporting - Not electronically submitting		Q14e_5	CDRNES
e. Clinical data registry reporting - Do not know		Q14e_6	CDRDK
f. Electronic reportable laboratory result reporting - Directly from Electronic health record		Q14f_1	ERLEHR
f. Electronic reportable laboratory result reporting - Health information exchange		Q14f_2	# ERLHIE #
f. Electronic reportable laboratory result reporting - Portal submission		Q14f_3	ERLPS
f. Electronic reportable laboratory result reporting - Flat files		Q14f_4	ERLFF
f. Electronic reportable laboratory result reporting - Not electronically submitting		Q14f_5	ERLNES
f. Electronic reportable laboratory result reporting - Do not know		Q14f_6	ERLDK
g. Antimicrobial use and resistance (AUR) - Directly from Electronic health record		Q14g_1	AMICEHR
g. Antimicrobial use and resistance (AUR) - Health information exchange		Q14g_2	# AMICHIE #
g. Antimicrobial use and resistance (AUR) - Portal submission		Q14g_3	AMICPS
g. Antimicrobial use and resistance (AUR) - Flat files		Q14g_4	AMICFF
g. Antimicrobial use and resistance (AUR) - Not electronically submitting		Q14g_5	AMICNES
g. Antimicrobial use and resistance (AUR) - Do not know		Q14g_6	AMICDK
h. Hospital capacity and utilization of medical supplies - Directly from Electronic health record		Q14h_1	HCUEHR
h. Hospital capacity and utilization of medical supplies - Health information exchange		Q14h_2	# HCUHIE #
h. Hospital capacity and utilization of medical supplies - Portal submission		Q14h_3	HCUPS
h. Hospital capacity and utilization of medical supplies - Flat files		Q14h_4	HCUFF
h. Hospital capacity and utilization of medical supplies - Not electronically submitting		Q14h_5	HCUNES
h. Hospital capacity and utilization of medical supplies - Do not know		Q14h_6	HCUDK


*Q15 - SSRSTRDT; IRRSTRDT; ECRSTRDT; PHRSTRDT; CDRSTRDT; ERLSTRDT; AURSTRDT; HCUSTRDT*?For each type of reporting does hospital use automated, manual or mix (binaryish; 1 = fully or primarily automated; 2 = mix or automated and manual process; 3 = fully or primarily manual; 4 = do not know)
a. Syndromic surveillance reporting - Public Health Reporting Type		Q15_A	SSRSTRDT
b. Immunization registry reporting- Public Health Reporting Type		Q15_B	IRRSTRDT
c. Electronic case reporting - Public Health Reporting Type		Q15_C	ECRSTRDT
d. Public health registry reporting - Public Health Reporting Type		Q15_D	PHRSTRDT
e. Clinical data registry reporting - Public Health Reporting Type		Q15_E	CDRSTRDT
f. Electronic reportable laboratory result reporting - Public Health Reporting Type		Q15_F	ERLSTRDT
g. Antimicrobial use and resistance (AUR) - Public Health Reporting Type		Q15_G	AURSTRDT
h. Hospital capacity and utilization of medical supplies - Public Health Reporting Type		Q15_H	HCUSTRDT


*Q16 - lots of variable - re-download if want to use*?What challenges has hospital experienced in past year when sending information electronically to PH? (categorical broken down by type of report)

*Q18 - PREDML; PREDOTH*Do you use predictive models/ML (categorical/binary - 1 = Yes; 2 = No; 3 = Do Not Know)
a. Machine Learning		Q18_a	PREDML
b. Other non-machine learning predictive models (e.g. APACHE IV)		Q18_b	PREDOTH
now coded as 0=no, 1=yes, 2=dnk


*Q19 - AITRAJ; AIRFOL; AIMHEA; AIRECT; AIBILL; AISCHE; AIPOTH; AIPOTH_DESC; AICLOTH; AICLOTH_DESC; AINOTA; AIDNK* For which do you applied ML or predictive (categorical - 1 = Yes; 0 = No)
a. Predicting health trajectories or risks for inpatients		Q19_A	AITRAJ
b. Identify high risk outpatients to inform follow-up care		Q19_B	AIRFOL
c. Monitor health		Q19_C	AIMHEA
d. Recommend treatments		Q19_D	AIRECT
e. Simplify or automate billing procedures		Q19_E	AIBILL
f. Facilitate scheduling 		Q19_F	AISCHE

g. Other (operational process optimization)		Q19_G	AIPOTH
Other description (operational process optimization)		Q19_G_OTH	AIPOTH_DESC
h. Other (clinical use cases)		Q19_H	AICLOTH
Other description (other clinical use cases)		Q19_H_OTH	AICLOTH_DESC

I. None of the above (machine learning application)		Q19_I	AINOTA
j. Do not know (machine learning application)		Q19_J	AIDNK

*HIE PARTICIPATION VARIABLE* +1 for how many HIE they participate in. +1 for is they use the HIE to exchange any of the ph data (is there a 1 in question 14 variables). Max score could be 14. Min score could be 0

*AI Composite Variable* Do they have a high use of AI? +1 for each activity apply AI toward in question 19, except none and do not know. min 0; max 8

*Final Variables*
1. AI Composite Score (continuous out of 8)
2. HIE Participation Score (continuous out of 14)
3. Number of Beds (continuous)
4. Ownership type (categorical)
5. Primary EMR (categorical)
6. State (categorical)

# Required
*Questions*: What hospital characteristics (state, bed size, ownership, EHR vendor, ML use?) predict active HIE participation? 
Does HIE participation predict more AI use?
Does more AI use predict more automated PH reporting? Does it interact with HIE paticipation
Do more people with Epic participate in HIE because of Cosmos?
Do a different type of ownership class more likely to participate in HIE? 
Are states with HIE more likley to participate?

*main interest is in hie participation, try to keep questions pointed that way.*
*Scatterplot*: participation vs bed total
*Boxplot*: participation in HIE by state; participation in state by type of ownerhsip
*Quantile*: multiple regression
*Histogram*: multiple regression
*T-test*: difference in HIE participation between ownership type? difference in HIE participation between ehr (cerner vs epic - epic vs not bc of cosmos); difference in hie based on automated ph use?
*Correlation*: bed total vs HIE participation? As hospitals are bigger are they more likely to participate in HIE
*Bootstrap of T-test or correlation*: 
*Permutation*: Is there a difference in owenship type and hie participation; if not, I can shuffle the tags of ownership and get the same statistic
*Multiple Regression*: does bed type, ownership type, primary emr, AI usage level predict HIE participation level
*ANCOVA*:
*Binary Logistic Regression*: can you predict if a hospital is or is not part of HIE
*Interaction plot*:

6) ONE OF THE FOLLOWING TECHNIQUES – ANOVA (with some form of multiple comparisons and preferably with a non-parametric analogue for comparison or Welch’s ANOVA – include discussion of interaction plots and interaction terms if two-way), ANCOVA, Binary Logistic Regression, Multinomial Logistic Regression

# Questions to ask office hours
0. What should i do for 6
1. Do the groups in t test need to be equal - ie ownership or primary EHR 
2. does there need to be a normal distribution for categorical (ie Epic leads the emr primary by a lot)
3. is it okay to do log bed total? do i need to do it?
5. does it matter that the distribution of hospisals by state is right skewed? One state has 300 hospitals
4. do you think i should keep all the graphs that I do have? ie boxploy, bargraph, and histogram for state hospital distirbution 
5. Are we allowed to use JDRSs source for residual plots?

are there any other cool tables or interactiosn you think i should look at



```{r}
data <- read.csv("AHA - Query.csv")
dim(data) # 25899, 28
names(data) 
head(data)
```

```{r}
sum(is.na(data$bdtot)) # 26 - bed total

sum(is.na(data$mcntrl)) # 2 - control/ownership

sum(is.na(data$piemro)) # 0 - all values present - primary inpatient EHR vendor

sum(is.na(data$nm1oth)) # 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name1 Other

sum(is.na(data$rhio_lvl1)) # 878 missing  Level of HIE or HIO participation

sum(is.na(data$hie_noex)) # - Do not exhange (HIE Vendor)
nrow(data) - sum(is.na(data$hie_noex)) # 356 values included

sum(is.na(data$hie_nodis)) # (hie_nodis) Prefer not to disclose (HIE Vendor
nrow(data) - sum(is.na(data$hie_nodis)) # 98 values included
# can we assume those who did not say prefer not to disclose or do not exchange have a HIE?
# maybe I include all answers and them manually bunch all together - do only for one year

sum(is.na(data$emoreg)) #  Regional, state, or local health information exchange organization (HIE/HIO). Not local, proprietary, enterprise network)
nrow(data) - sum(is.na(data$emoreg)) # 11687 values included

sum(is.na(data$nmref1)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 1
nrow(data) - sum(is.na(data$nmref1)) # 5620 values included 

sum(is.na(data$year1)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 1
nrow(data) - sum(is.na(data$year1)) # 5606 values included

sum(is.na(data$nmref2))#  Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 2
nrow(data) - sum(is.na(data$nmref2)) #1681 values included

sum(is.na(data$year2)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 2
nrow(data) - sum(is.na(data$year2)) # 2046 values included

sum(is.na(data$nm2oth))# 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name2 Other

sum(is.na(data$nmref3)) # Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name/reference number 3
nrow(data) - sum(is.na(data$nmref3)) # 832 values included

sum(is.na(data$year3)) #Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO.-First year your hospital participated 3
nrow(data) - sum(is.na(data$year3)) #1160 values included 

sum(is.na(data$nm3oth)) # 0 - all values present - Please indicate which state, regional and/or local HIE/HIO(s) with whom your hospital participates and actively exchanges data as well as the first year that your hospitals actively participated in this HIE/HIO. - Name3 Other

sum(is.na(data$ssrshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Syndromic surveillance reporting
nrow(data) - sum(is.na(data$ssrshie)) # 2060 values included 
sum(is.na(data$irrshie)) #  If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Immunization registry reporting
nrow(data) - sum(is.na(data$irrshie)) # 2037 values included

sum(is.na(data$ecrshie)) # f you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Electronic case reporting
nrow(data) - sum(is.na(data$ecrshie)) #2006 values included

sum(is.na(data$phrshie)) #  If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Public health registry reporting
nrow(data) - sum(is.na(data$phrshie)) # 2005 values included 
sum(is.na(data$cdrshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Clinical data registry reporting
nrow(data) - sum(is.na(data$cdrshie)) # 2003 values included 
sum(is.na(data$erlshie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Electronic reportable laboratory result reporting
nrow(data) - sum(is.na(data$erlshie)) # 2043 values included

sum(is.na(data$hcushie)) # If you use an HIE to submit data for public health reporting activities, does the HIE perform value added services during submission? Hospital capacity and utilization of medical supplies
nrow(data) - sum(is.na(data$hcushie)) # 1970 values included

sum(is.na(data$hiechrge)) # HIE charges your hospital additionally to submit data for public health reporting activities.
nrow(data) - sum(is.na(data$hiechrge)) # 2643 values included 

```


```{r}
data <- read.csv("AHA Data - Query 2.csv")
dim(data)
names(data)
```

