---
title: "Homework 08 One Way ANOVA"
output:
  word_document: default
  html_document: default
  pdf_document: default
subtitle: Due by 11:59pm, Tuesday, 4.14.26
author: S&DS 2300/5300/ENV 757
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
The CSV file `HindiFilm.csv` which you can get [HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/HindiFilm.csv) contains the 1698 Hindi language movies that released in India across a 13 year period (2005-2017)  (thanks to Premkumar, Prashant (2020), “Bollywood Movies data”, Mendeley Data, V1, doi: 10.17632/3c57btcxy9.1).  Here are the variables:

* Name
* Period (normal or holiday)
* Remake (Yes/No)
* Franchise (Yes/No)
* Genre
* New_Actor (Yes/No)
* New_Director (Yes/No)
* New_Musician (Yes/No - i.e. new music director)
* Lead (Lead actor name)
* Director (Name)
* Musician (Music Director name)
* Screens (how many screens was it shown on)
* Revenue (Indian Rupees)
* Budget (Indian Rupees - usual shorthand is INR)



**1)  One Way ANOVA of percent return for movies.** *(75 pts - 6 pts each except part 1.12 which is 9 pts)*

1.1)  Read the data into an object called `movie` (do NOT use the option `as.is = TRUE`).   Update this object so that is only contains the columns `Budget`, `Revenue`, `Genre`, and `Name`.  In addition, retain only rows that have complete data for these four columns.

Next, create a new column called `pctReturn` that calculates the percentage return (have this be a percentage, NOT a fraction).

Make a table of `Genre` and sort this from high to low.  Show these results.

Update `movie` so that it only contains movies in the following genres : "masala", "drama", "fantasy", "love_story", "action".

As per usual, look at the first few rows of the data, get the dimension of the final dataset.

```{r}
# read data in
movie <- read.csv("https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/HindiFilm.csv")

# only include budget, revenue, genre, and name; only complete data
movie <- na.omit(movie[ , c("Budget", "Revenue", "Genre", "Name")])

# create column pctResturn; percentage return as percent
movie$pctReturn <- ((movie$Revenue - movie$Budget)/movie$Budget)*100

# table of genre sorted from high to low
sort(table(movie$Genre), decreasing = TRUE)

# movie genres only masala, drama, fantasy, love_story, action
movie <- movie[movie$Genre %in% c("masala", "drama", "fantasy", "love_story", "action"), ]

# examine data and get dimensions
head(movie)
dim(movie)
range(movie$pctReturn)
```


1.2)  Make a boxplot of `pctReturn` by `Genre`.   Superimpose a red dot at the mean percent return for each Genre; also print this mean rounded to the nearest percent. In addition, add a horizontal dotted blue line at 100 (this represents the place where revenue = budget).  Discuss what you observe.

```{r}
# boxplot of pctReturn by genre with red dot at mean percent return for each genre and horizontal bluye dotted line at 100
# CHECK: is this plot right?
boxplot(movie$pctReturn ~ movie$Genre,
        col = "steelblue",
        ylab = "", 
        xlab = "Genre Type",
        cex.axis = .8,
        ylim = c(range(movie$pctReturn)[1] - 20000, range(movie$pctReturn)[2]),
        las = 1,
        main = "Boxplot of Percent Returned by Genre")
means <- tapply(movie$pctReturn, movie$Genre, mean)
points(means, col = "red3", pch = 18)
text(x = c(1:5), y = means - 10500, labels = paste(round(means, 2), "%"), cex = .8, font = 1)
abline(h = 100, col = "blue3", lty = 3, lwd = 2)
```


*The IQR and whiskers (so essentially the whole box plot) is very squished making it practically unreadable. We are able the appreciate the outliers and how far away some points are, meaning they generated a lot of revenue relative to their budget. From the means, we see most films are lucrative and making more money on average then took to produce with fantasty, drama and love story especially make high revenue margins. Masala movie on average los money; althought meanness is easily pulled by outliers, which we see are very evident here and to varying degress across genres (meaning we should assume their variances will be different), so it's not a great measure to examine in this case.*

1.3)  Make a new variable on 'movie' called `log10pct` that is log base 10 of percent return (the function you want is `log10()`).  Repeat part 1.2, but modify the horizontal line so that it is at the correct value (when rounding mean log10(pct return), round to 1 decimal place).  Discuss what you observe.

```{r, warning=FALSE}
# range includes 0s add noise to all to have no 0 value 
# CHECK: why are we not doing logit here
# CHECK: Is it okay if there are NANs
movie$log10pct <- log10(movie$pctReturn + 0.05)

# repeat part 2 with horizontal line at correct mean value
boxplot(movie$log10pct ~ movie$Genre,
        col = "steelblue",
        ylab = "", 
        xlab = "Genre Type",
        cex.axis = .8,
        las = 1,
        main = "Boxplot of Percent Returned by Genre")
means <- tapply(movie$log10pct, movie$Genre, mean, na.rm = TRUE)
points(means, col = "red3", pch = 18)
text(x = c(1:5), y = means + .3, labels = paste(round(means, 2), "%"), cex = .8, font =2)
abline(h = round(mean(means), 1), col = "blue3", lty = 3, lwd = 2)
```


*Using the log values, we can appreciate the box plots a lot better. Here, we can see most genres hover around the break even line, especially action, and masala actually is completely below it. There are lots of outliers to the negative side in drama and the variances look a little more similar for the first 4 genres.*

1.4)  Make a `stripplot` of log10pct by Genre.   Which plot do you find more instructive - the boxplot or the strip plot (no right answer here, but give a one sentence reason for your opinion)?

```{r}
# make stripplot of log10pct by genre
#install.packages('lattice')
library(lattice)
stripplot(movie$log10pct ~ movie$Genre,
          col = "forestgreen",
          ylab = "",
          xlab = "Genre Type",
          main = "Percent Returned on Log Scale by Genre Type")
```


*Typically, I like a box plot because of the extra information it provides, but for this data, I actually prefer the stripplot. The IQR box and whiskers aren't adding much additional information, and I appreicate that with the stripplot the range is very evident and appreciatable.*

1.5)  Just using the visual information provided by the boxplot of log10pct, do you think that the distribution of observations inside each Genre is approximately normal?   Do you think the standard deviation is approximately the same in each Genre?  (Remember that ANOVA assumes we have a normal distribution inside each Genre and that the standard deviation is the same across groups.)

*I do  think the distribution of observations inside each Genre is normal because the whiskers and IQR are relatively equal.*
*No, I do not think the standard deviation is approximately the same in each Genre - their ranges are very inconsistent, some are long and some are short. This is especially appreciatable in the strippplot.*

1.6)  Calculate the sample standard deviation of log10pct for each Genre   Calculate the ratio of largest to smallest sample standard deviation.   Is it reasonable to assume that the variances are the same across Genres?

```{r}
# sample standard deviation of log10pct for each genre
(sds <- tapply(movie$log10pct, movie$Genre, sd, na.rm = TRUE))

# calculate ratio of largest to smallest sample sd
round(max(sds)/min(sds), 1)
```


*Yes, it is reasonable to assume the variance is the same across genres, because the ratio is less than 2.*

1.7)  Use the `aov()` function to compare mean log10pct between Genres.  Save your results to an object called `aov1`.  Get summary information for `aov1`.  Is the mean log10pct return statistically significantly different between Genres?  In addition, confirm that the degrees of freedom reported by the test are what you expect (and write a sentence about this).

```{r}
# use aov() function to compare mean log10pct between genres in aov1
aov1 <- aov(log10pct ~ Genre, data = movie)

# summary of aov1
summary(aov1)

# number of observations without na
length(aov1$fitted.values)

```


*Yes, the p-value is essentially 0, so it is statistically significant, meaning the means between genres is different.*
*Genre degree of freedom (5 genres - 1 = 4) makes sense; residual df (692 - 5 = 687) makes sense*

1.8)  Fit log10pct return based on Genre as a regression model WITHOUT an intercept.  Save the results to an object called `mod1`.  Following the example code provided in class, calculated confidence intervals for the mean percent return for each Genre and report results to two decimal places.  Then make a plot of the resulting confidence intervals using the `plotCI()` function in the `plotrix` library.

You'll notice that some intervals overlap while others do not.   What should you conclude from this?

```{r}
# fit log10pct return based on genre as a regression model without intercept in mod1
mod1 <- lm(movie$log10pct ~ movie$Genre -1)

# calculate confidence interval for mean percent return for each genre to two decimal places
CIs <- confint(mod1)
coefs <- coef(mod1)

# plot of resulting ci using plotCI() in plotrix
#install.packages('plotrix')
library(plotrix)
par(mar = c(5, 8, 4, 2))
plotCI(coefs, 1:(length(coefs)), ui = CIs[,2], li = CIs[,1], axes = FALSE, err = "x", ylab = "", xlab = "Mean log(Percent Revenue) and 95% CI", main = "Mean and CI's of Percent Revenue in log", lwd = 2, col = "blue")
axis(side=1)
axis(side = 2, at = 1:(length(coefs)), label = levels(as.factor(movie$Genre)), las = 2)
```


*We cannot compare between groups with this plot. But, we can conclude that some intervals pass 0 while others do not. For love story, fantasy, drama, and action there is evidence that they are different. CHECK*

1.9)  Use the `pairwise.t.test()` function to calculate Holm's correction for comparing pairs of means.  Which pairs of Genres have statistically significantly different percent returns (use alpha = 0.05)?

```{r}
# use pairwise.t.test() to calculate Holms's correction
pairwise.t.test(movie$log10pct, movie$Genre)
```


*drama - action; love story - action; masala - action; masala - drama; masala - fantasy; masala - love story*


1.10)  Calculate Tukey simultaneous 95% confidence intervals for differences in mean percent return using the `TukeyHSD()` function.  Plot the resulting confidence intervals.  Do you reach the same conclusions as when using the Holm correction?

```{r}
# Tukey simultaneous 95% CI for differences in mean percent return
TukeyHSD(aov1)

# plot resulting CI
par(mar=c(5, 8, 4, 2))
plot(TukeyHSD(aov1), 
     las = 1)
```


*drama - action; love story - action; masala - action; masala - drama; masala - fantasy; masala - love story. Yes you do reach the same conclusino as using the Holm correcton.*


1.11)  Create residual plots appropriate for `aov1`.  Does it appear that the residuals have an approximately normal distribution?  Is this good or bad?  Comment in a sentence or two about the plot of fits versus residuals.

```{r}
# residual plot for aov1
#Get a function for doing matrix plots and also for regression residuals - from lecture 19
source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt")
myResPlots(aov1, "Log Percent Return")
```


*They almost have a normal distribution except a long left tail. A normal distribution is good because ANOVA assumes the data is normal. The plot of fitted vs residuals does have some heteroskedasticity. Because the predicted values are categorical, we expect them to be in vertical lines like they are, but we would want their lengths to be more similar to each other.*


1.12)  Just to see what would have happened if we didn't use a log10 transformation, repeat parts 1.7 through 1.11 on the original percent return data (i.e. not log10).  Just use one block of code (literally, just copy your code, replace `log10pct` with `pctReturn`). 

Write about three sentences commenting on the fit of this model vs. the fit of the previous model.  Specifically, comment on whether any pairs of groups changed significance, the overall model significance, and the residual plots.

```{r}
# repeat 1.7
#use aov() function to compare mean log10pct between genres in aov1
aov1 <- aov(pctReturn ~ Genre, data = movie)
# summary of aov1
summary(aov1)
# number of observations without na
length(aov1$fitted.values)

# 1.8
# fit percent return based on genre as a regression model without intercept in mod1
mod1 <- lm(movie$pctReturn ~ movie$Genre -1)
# calculate confidence interval for mean percent return for each genre to two decimal places
CIs <- confint(mod1)
coefs <- coef(mod1)
# plot of resulting ci using plotCI() in plotrix
#install.packages('plotrix')
library(plotrix)
par(mar = c(5, 8, 4, 2))
plotCI(coefs, 1:(length(coefs)), ui = CIs[,2], li = CIs[,1], axes = FALSE, err = "x", ylab = "", xlab = "Mean Percent Revenue and 95% CI", main = "Mean and CI's of Percent Revenue", lwd = 2, col = "blue")
axis(side=1)
axis(side = 2, at = 1:(length(coefs)), label = levels(as.factor(movie$Genre)), las = 2)

# 1.9
# use pairwise.t.test() to calculate Holms's correction
pairwise.t.test(movie$pctReturn, movie$Genre)

# 1.10
# Tukey simultaneous 95% CI for differences in mean percent return
TukeyHSD(aov1)

# plot resulting CI
par(mar=c(5, 8, 4, 2))
plot(TukeyHSD(aov1), 
     las = 1)

# 1.11
myResPlots(aov1, "Percent Return")
```


*Using the raw values, no interactions are signficant and the data appears to be veryyyy right skewed.*


**2)  Non-parametric tests.** *(25 pts - 5pts each)*


2.1)  Perform a Bartlett test to see whether variances of percent return are the same across Genres.  Then, repeat for `log10pct`.  Write one sentence about what you conclude.

```{r}
# Bartlett test if variances of percent return are same across genre
bartlett.test(movie$pctReturn, movie$Genre)

# repeat Bartlett test for log10pct
bartlett.test(movie$log10pct, movie$Genre)
```

*For the raw values, the p-value is nearly zero, so we can reject the null hypothesis and conclude that the variances are different between groups. For the log values however, there is a high p-value so we assume the variance is the same between the groups - this indicates we would rather use the log variable. Bartlett assumes normal distribution which is more aligned with the log data distribution.*


2.2)  Perform a Levene test to see whether variances of percent return are the same across Genres.  Then, repeat for `transpct`.  Write one sentence about what you observe (be sure to review the characteristics of Levene's test).

```{r, warning = FALSE}
# Levene test of variances of percent return are same across genre
library(car)
leveneTest(movie$pctReturn, movie$Genre)

# repeat for transpct
# CHECK: is this log10pct
leveneTest(movie$log10pct, movie$Genre)

```


*Both p values are not statistically significant therefore we retain the null hypothesis that the variances are the same between groups for both the raw variables and the log variables. The Levene's test does not assume the observations are from a normal distribution so it can be applied to the raw values, but it is less powerful which is how we conclude the variations are the same for all groups within both variables altough we know it is not true for the raw values.*

2.3)  Perform Welch's ANOVA on the original data which you will recall assumes unequal variances but which assumes normal distributions in each group  (i.e. compare `pctReturn` by `Genre`).  Compare results to regular one-way ANOVA.  Is Welch's ANOVA a good choice here?  

```{r}
# Welch's ANOVA on original data (percent return by genre)
oneway.test(pctReturn ~ Genre, data = movie)

# One Way ANOVA
summary.aov(aov(pctReturn ~ Genre, data = movie))
```

# CHECK should I be looking at F or p
*In the regular one-way ANOVA, the p value is not significant at an alpha of 0.05 meaning the groups are not different. In the Welch's test, the p-value is very close to 0 indicating the groups are different, but it assumes a normal distribution which we know a not true and thus would not be a good choice here.*


2.4)  Perform a non-parametric Kruskal Wallis test which recall makes NO assumptions about equal variances or about normality.  Compare results to regular one-way ANOVA.  Is Kruskal Wallis a good choice here?  

```{r}
# non-parametric Kruskal Wallis test 
kruskal.test(pctReturn ~ Genre, data = movie)

# One Way ANOVA
summary.aov(aov(pctReturn ~ Genre, data = movie))
```


*Compare results to regular one-way ANOVA.*  
*Is Kruskal Wallis a good choice here?  *

2.5)  Repeat 2.3) and 2.4) on the log10 scale.  Discuss your results and compare results to regular ANOVA.

```{r}
# repeat 2.3 run Welch's test on log10 scale
oneway.test(log10pct ~ Genre, data = movie)

# repeat 2.4 run Kruskal Wallis test on log10scale
kruskal.test(log10pct ~ Genre, data = movie)

# compare to regular ANOVA
summary.aov(aov(log10pct ~ Genre, data = movie))

```


*Using log 10, all tests return significant at essentially any alpha level. This would indicate that the groups means are not the same, which is also what we see in our regular one-way ANOVA.*

3)  Pseudocode
You are working with a dataset of average daily tourism spending for five different types of visitor experiences in Connecticut: coastal, city/cultural, outdoor recreation, culinary, and family-oriented. Each observation represents a week’s worth of spending data collected over the past three years. The dataset also includes information on total visitors, number of local events, and whether the data week included a holiday period. Your task is to determine whether average spending per visitor differs among experience types and to check whether assumptions of equal variances and normality of residuals are met.

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

