---
title: "Homework 05 Functions and Permutation Tests"
output:
  word_document: 
     fig_height: 5
     fig_width : 7
  pdf_document: 
     fig_height: 5
     fig_width : 7
  html_document: default
subtitle: Due by 11:59pm, Tuesday, 3.3.26
author: S&DS 2300/5300/ENV 757
---



```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Questions 1 and 2 use data from both  2017 and 2018 New Haven Road Races - in particular, we look at 5k run times. You can get data for [2018 HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/NHRR2018.csv) and for [2017 HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/NHRR2017.csv).

**1) Function for Data Cleaning** *(20 points)*

(1.1) *(5 pts)* Load in both .csv files into objects called `nh2017` and `nh2018`. Use `head()`, `names()`, and `str()` to check if both datasets have the same variable names and the same format (i.e does each variable have the same format in each dataset).  Comment on what you observe.

```{r}
nh2017 <- read.csv('https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/NHRR2017.csv')
#head(nh2017)
str(nh2017)
names(nh2017)

nh2018 <- read.csv('https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/NHRR2018.csv')
head(nh2018)
str(nh2018)
names(nh2018)
```


*Both datasets have the same variables (No., Name, City, Div, Time, Pace, Nettime) with the same format. There are more observations in 2017, so they had more runners. *

(1.2)  *(15 pts)*  Since the two datasets seem to have the same structure, we can write a function that creates new variables in each dataset. This function will be called `cleanNHData()`. As a first step, I've already included code to load the `lubridate` package and define a function called `convertTimes()` similar to that we used in Class 10.

I've started the outline of the function below.  Your job is to follow the exact process we used in class 9 to clean the 2018 data.  You need to replace each comment line in the `cleanNHData()` function with the code that will perform this task.   You literally just need to find the relevant line in the class code and put this into the `cleanNHData()` function.  The one exception is a new line you'll need to write that deletes rows where `Name` is missing (i.e. equal to "")

Then, run the function on `nh2017` and `nh2018` to replace each of these datasets with the cleaned up version of themselves.


```{r, warning = FALSE}
library(lubridate)
convertTimes <- function(v) {
  hourplus <- nchar(v) == 7
  wrongformat <- nchar(v) == 8
  outtimes <- ms(v)
  if (sum(hourplus) > 0) { # if there is at least 1 time that exceeds 1 hr
    outtimes[hourplus] <- hms(v[hourplus])
  }
  if (sum(wrongformat) > 0) { # if there is at least 1 time in wrong format
    outtimes[wrongformat] <- ms(substr(v[wrongformat],1,5))
  }
  outtimes <- as.numeric(outtimes)/60
  return(outtimes)
}

cleanNHData <- function(data) {
  data$Div[data$Div == ""] <- NA 
  data$Gender <- substr(data$Div, 1, 1)
  data$AgeGrp <- substr(data$Div, 2, nchar(data$Div))
  data$Nettime_min <- convertTimes(data$Nettime)
  data$Time_min <- convertTimes(data$Time)
  data$Pace_min <- convertTimes(data$Pace)
  
  #Replace dataset with same dataset such that Name is not equal to ""
  data <- data[data$Name != "", ]                      
  
  #Return the dataset
  return(data)                                         
}

#run cleanNHData on nh2018 and nh2017 and replace these with the cleaned up versions of themselves
nh2017 <- cleanNHData(nh2017)
nh2018 <- cleanNHData(nh2018)
names(nh2018)
names(nh2017)
head(nh2017)
head(nh2018)
```


**2) Repeat Runners Dataset** *(35 points)*

We now create a dataset that looks at times of runners who ran in both 2018 and 2017.  

(2.1)  *(5 pts)* We'll have problems if we have instances of two runners having the same name.   A crude fix is to delete the second occurance of anyone with a duplicate name.

Run the code below to see how the function `duplicated()` works:

```{r}

duplicated(c("cat", "cat", "dog", "llama"))

```

Esentially, this returns a vector that is `FALSE` if an observation value is the first occurrence of this value and `TRUE` when a value has been seen before.   

To merge our two datasets, we need to start with unique `Name` values in each dataset.  Using the `duplicated()` function, create two new dataframes called `nh2018Unq` and `nh2017Unq` so that each only retains observations for the first occurence of each value of `Name` (if you use the `!` operator, this is two short lines of code).  

Get the dimensions of each of the four relevant dataframes.   How many observations were eliminated from each year?

```{r}
# find unique names in each dataset
nh2017Unq <- nh2017[!duplicated(nh2017$Name), ]
nh2018Unq <- nh2018[!duplicated(nh2018$Name), ]

# dimensions of each dataframes
dim(nh2017)
dim(nh2017Unq)
dim(nh2018)
dim(nh2018Unq)

# how many observations eliminated 
dim(nh2017) - dim(nh2017Unq)
dim(nh2018) - dim(nh2018Unq)
```


*7 observations were eliminated from 2017. 45 were eliminated from 2018.*

(2.2)  *(5 pts)* Next, we need to get a list of names that occur in both datasets.  Run the code below to see how the `intersect()` function works.

```{r}
intersect(c("cat", "dog", "llama"), c("cat", "llama", "chincilla"))
```

Using the `intersect()` function, create an object called `repeatrunners` that is a list of names of people who ran in both years.  How many runners ran in both years?

```{r}
repeatrunners <- intersect(nh2017Unq$Name, nh2018Unq$Name)
length(repeatrunners)
```

*986 runners ran both years*

(2.3)  *(15 pts)* The code below will create a combined dataset called `nhcombined`.  Your job in this section is to write a one or two line comment above each line of code to describe what the line does.  You'll want to run each line, probably see what the result was, and in some cases use the help file for some functions to see what the function does (i.e. for the `merge()` function).  Make sure you remove `eval = FALSE` in the r chunk.


```{r}
# this creates a logical vector stored in w that shows if a runner in 2018 is one of the repeat runners 
w <- nh2018Unq$Name %in% repeatrunners 

# this created a df that has columns filled with the repeat runners name gender and nettime in 2018.
nhcombined <- data.frame(Name = nh2018Unq$Name[w],
                         Gender = nh2018Unq$Gender[w],
                         Nettime_2018 = nh2018Unq$Nettime_min[w])

# this merges the df nhcmobined and nh2017uniq's name and nettime columns. it adds the nettime 
nhcombined <- merge(nhcombined, nh2017Unq[ ,c("Name", "Nettime_min")])

# This removes any runner with NA as their gender and rewrites it into nhcombined
nhcombined <- nhcombined[!is.na(nhcombined$Gender),]

# this changes the name of column 4 in nhcombined to Nettime_2017
colnames(nhcombined)[4] <- "Nettime_2017"

# this gets the dimensions of the final df - 985 runners 
dim(nhcombined)

# shows the first 6 details of the duplicate runners 
head(nhcombined)

```

(2.4)  *(5 pts)* Create a new variable in the data frame `nhcombined` called `improvement` that is the improvement in run time from 2017 to 2018 (a positive number here should indicate an improvement,a negative number means they did worse in 2018). Get summary statistics for `nhcombined`.  Then make a histogram of `improvement`.  Comment on the summary statistics and what you observe in the histogram.


```{r}
# create new improvement 
nhcombined$improvement <- nhcombined$Nettime_2017 - nhcombined$Nettime_2018
  
#summary of nhcombined
summary(nhcombined)

# histogram of improvement 
hist(nhcombined$improvement,
     main = "Histogram of Change in Time \n between 2017 and 2018 Duplicate Runners",
     breaks = 75, # increase to better see data distribution
     xlab = "Change in Time")
```


*The majority of runners did not see a large change in their time but there was overall improvement seen in runners who ran in both 2017 and 2018. The mean improvement time was -1.11 and the median was -0.93. There are some presumed outliers (a boxplot would better annalyze this but we can see the x axis of histogram is stretched), but the IQR of improvement - which is at least 50% of the data is between -2.6 and 0.5 minute changes between the years so most people stayed consistent. We see this in the histogram because the majority of the data stays at or around 0.*

(2.5)  *(5 pts)* You'll notice a few extreme values (i.e. people got amazingly better or worse).   Print the rows of `nhcombined` that had improvement times of more than 50 in absolute value.  Update the `nhcombined` dataframe to exclude these rows and make the histogram again.

```{r}
# print extreme values 
print(nhcombined[abs(nhcombined$improvement) > 50, ])

# update nhcombine to exclude extreme values
nhcombined <- nhcombined[abs(nhcombined$improvement) <= 50, ]

# histogram of nhcombined improvement times again without extreme values 
hist(nhcombined$improvement,
     main = "Histogram of Change in Time \n between 2017 and 2018 Duplicate Runners \n (without Extreme Times)",
     breaks = 75, # Maintain same break
     xlab = "Change in Time")
```


**3) Changes in Measles Vaccination Rates in the Past 8 Years** *(45 pts)*

Question 3 uses data from from the 2016 and 2024 World Bank datasets. You can get data for [2016 HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/WB.2016.csv) and for [2024 HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/WB_2024.csv).

(3.1)  *(10 pts)* Read in the datasets.  Get and show the names on each dataset.  Confirm that the dimensions are the same between datasets.
Then, modify each dataset so that it only contains "Country", "Measles", and "GNI".  

Following the example in question 2, combine these datasets together. You'll need to rename "Measles" and "GNI" in each dataset before combining based on Country (something like "Measles_24", "Measles_16", etc.) In the combined dataset, remove any observations that are missing for either Measles variable or for GNI in 2024.

Create a new factor variable that identifies countries as having GNI in 2024 greater than 8000 or less than or equal to 8000.

Finally, calculate a variable that is the change in Measles vaccination rates (2024 minus 2016) per country.

Show the first 10 rows of the final dataset.

```{r}
# read in data sets
wb2016 <- read.csv('https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/WB.2016.csv')
wb2024 <- read.csv('https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/WB_2024.csv')

# show names of datasets
names(wb2016)
names(wb2024)

# confirm dimensions are same between datasets
dim(wb2016) == dim(wb2024) # only rows are same; different number of columns 
dim(wb2016) # 217 - 29
dim(wb2024) # 217 - 18 
# the amount of countries investigated are the same but in 2024, they looked at less variables (ie less columns)

# modify each to only contain Country, Measles, and GNI
wb2016 <- wb2016[ , c("Country", "Measles", "GNI")]
wb2024 <- wb2024[ , c("Country", "Measles", "GNI")]

# confirm both only have country, measles, and gni
names(wb2016)
names(wb2024)

# rename variables to be for each year
names(wb2016) <- c("Country_2016", "Measles_2016", "GNI_2016")
names(wb2024) <- c("Country_2024", "Measles_2024", "GNI_2024")

# combine datasets together with renamed variables 
# ensure all names are unique in each dataset
duplicated(wb2016$Country_2016) # no duplicates within 
duplicated(wb2024$Country_2024) # no duplicates

# ensure all countries appear in both 
setequal(wb2016$Country_2016, wb2024$Country_2024) #FALSE

# Countries in 2016 but not in 2024
setdiff(wb2016$Country_2016, wb2024$Country_2024)

# Countries in 2024 but not in 2016
setdiff(wb2024$Country_2024, wb2016$Country_2016)

# Czech Republic - Czechia; Swaziland - Eswatini; Korea, Dem. People\x92s Rep. - Korea, Dem People's Rep.; Macedonia, FYR - North Macedonia; Turkey - Tukiye; Vietnam - Viet Nam

# correct these country names to be aligned to what is in 2024
wb2016$Country_2016[wb2016$Country_2016 == "Czech Republic"] <- "Czechia"
wb2016$Country_2016[wb2016$Country_2016 == "Swaziland"] <- "Eswatini"
wb2016$Country_2016[wb2016$Country_2016 == "Korea, Dem. People\x92s Rep."] <- "Korea, Dem. People's Rep."
wb2016$Country_2016[wb2016$Country_2016 == "Macedonia, FYR"] <- "North Macedonia"
wb2016$Country_2016[wb2016$Country_2016 == "Turkey"] <- "Turkiye"
wb2016$Country_2016[wb2016$Country_2016 == "Vietnam"] <- "Viet Nam"

# recheck all are same
setequal(wb2016$Country_2016, wb2024$Country_2024) # TRUE!

# stores values of country in c of countries in 2016 that appear in 2024
c <- wb2016$Country_2016 %in% wb2024$Country_2024
wb2016$Country_2016 == wb2024$Country_2024

# create initial dataframe of 2016 values 
wb <- data.frame(Country = wb2016$Country_2016[c],
                 Measles_2016 = wb2016$Measles_2016[c],
                 GNI_2016 = wb2016$GNI_2016[c])

# merges in 2024 dataset, mathcing based on country names 
wb <- merge(wb, wb2024[ , c("Country_2024", "Measles_2024", "GNI_2024")],
            by.x = "Country", by.y = "Country_2024") # will only keep by.x, drop by.y since they match

# remove observations missing for Measles or GNI in 2024
wb <- wb[!is.na(wb$Measles_2024),]
wb <- wb[!is.na(wb$Measles_2016),]
wb <- wb[!is.na(wb$GNI_2024),]

# create new factor variable that identified countries with GNI_2024 > 8000 | <= 8000
wb$GNI_8000 <- as.factor(ifelse(wb$GNI_2024 > 8000, "Above 8000", "Below or Equal to 8000"))

# calculate change in measles vaccination rate per country
wb$Measles_change <- wb$Measles_2024 - wb$Measles_2016

# show first 10 rows of final dataset
head(wb, n=10)

```


(3.2)  *(10 pts)* Calculate and display summary statistics for the change in Measles vaccination rates overall.  Make a histogram of these changes and add a vertical line at the value which indicates no change. Discuss what you observe in a few sentences.

Make a side-by-side boxplot to see differences between the change in Measles vaccination rates between Countries with GNI < 8000 and countries with GNI > 8000.   Does there appear to be any difference between groups?  Comment both on center and spread.

```{r}
# calculate and display summary statistics for change in measles 
summary(wb$Measles_change)

# create histogram
hist(wb$Measles_change, 
     main = "Histogram of Rate Changes in Measles Vaccination \n Between 2024 and 2016",
     breaks = 20,
     xlab = "Changes in Measles Vaccination Rate",
     col= 'grey60')

# add vertical line at no change
abline(v = 0,
       lwd = 4,
       col = 'peru')

text(-1.5, 30, "No Rate Change", srt=90, col = 'white')

# side by side box plot of differences between change in Measles vaccination of countries on either side of 8000
boxplot(wb$Measles_change ~ wb$GNI_8000,
        main= "Boxplot of Measles Vaccination Rate Change \n Betweeen 2016-2014 in Countries GNI\n Above or Below 8000",
        ylab = "Change in Measles Vaccination Rate",
        xlab = "GNI Status", 
        col = "forestgreen")

# Extra
# confirm summary statistics of each 
by(wb$Measles_change, wb$GNI_8000, summary)
```


*Looking at the histogram, The majority of countries saw a decrease in measles vaccination rate. The mean and median was -3% and -1% decrease, respectively. There are some countries that had as much of a 32% increase in vaccinations but some also had a 45% decrease. The spread does look normal (maybe slightly left skewed considering the mean is less than the median).*
*Looking at the box plots, we see how the outliers and spread differs based on countries GNI. The countries with a GNI less than or equal to 8000 had just a slightly larger decrease in measles vaccination rate as the median for that box is lower. They also had a larger spread, both within the IQR and the whiskers and had more outliers that was an increased vaccination rate. The spread for the countries with a GNI above 8000 is much tighter, meaning overall the data was more condensed. There are several outliers, however; many that are a larger decrease than the rest of the data and one that is an increase. *

(3.3) *(10 pts)* Create a 95% bootstrap confidence interval for the mean change in measles vaccination rates among countries with GNI < 8000. Do the same for countries with a GNI > 8000. You don't need to make any histograms of your bootstrap results, and you don't need to use the `t.test()` function.  You also are not comparing the means of these two groups - you're getting separate intervals for each GNI group.  Display the intervals and discuss what you observe.


```{r}
# To make grading easier, please leave the following line of code in your assignment
set.seed(230)

# FILL IN REMAINING CODE
# create bootstrap for < 8000
mean_greater_8000 <- rep(NA, 10000)

for (i in 1:10000){
  s <- sample(wb$Measles_change[wb$GNI_8000 == "Above 8000"], sum(wb$GNI_8000 == "Above 8000"), replace = TRUE)
  mean_greater_8000[i] <- mean(s)
}

# get 95% confidence interval
qg <- quantile(mean_greater_8000, c(0.05, 0.95))
paste("95% CI Interval of Measles Vaccation Rates of Countries with GNI > 8000", round(qg[1], 2), "to", round(qg[2], 2))

# repeat bootstrap for > 8000
mean_less_8000 <- rep(NA, 10000)

for (i in 1:10000){
  s <- sample(wb$Measles_change[wb$GNI_8000 == "Below or Equal to 8000"], sum(wb$GNI_8000 == "Below or Equal to 8000"), replace = TRUE)
  mean_less_8000[i] <- mean(s)
}

# get 95% confidence interval
ql <- quantile(mean_less_8000, c(0.05, 0.95))
paste("95% CI Interval of Measles Vaccation Rates of Countries with GNI <= 8000", round(ql[1], 2), "to", round(ql[2], 2))

#Extra Check
# confirm p value with t-test 
t.test(wb$Measles_change[wb$GNI_8000 == "Above 8000"]) #-3 to -.6 p = 0.004
t.test(wb$Measles_change[wb$GNI_8000 == "Below or Equal to 8000"]) #-6 to -1 p = .0006

```

*From the bootstraping, we can see that the countries with GNI > 8000 have a smaller confidence interval while the countries with a GNI less than or equal to 8000 has a larger confident interval. Both confidence intervals are completely negative and do not include 0, meaning we potentially can reject our null hypothesis and accept that is a change in measles vaccination rate with the rate declining. This would be true for both groups, although the countries with a GNI less than or equal to 8000 is potentially declining at a steeper rate.*

(3.4) *(15 pts)* Using a permutation test, examine whether there a significant difference in the **MEDIAN** change in vaccination rates between high GNI countries and low GNI countries (calculate as high - low). Use a significance level of 0.01. Be sure to state (in words is fine) the null and alternative hypotheses, and discuss your conclusion.  Be sure to include a histogram of results and add a vertical line that shows that observed difference in medians (see example in code from class).

```{r}
# To make grading easier, please leave the following line of code in your assignment
set.seed(230)

# FILL IN REMAINING CODE
# imput low as "Below or Equal to"
levels(wb$GNI_8000)
levels(wb$GNI_8000)[levels(wb$GNI_8000) == "Below or Equal to 8000"] <- "low"

# imput high as "Above 8000"
levels(wb$GNI_8000)[levels(wb$GNI_8000) == "Above 8000"] <- "high"
levels(wb$GNI_8000)

# determine difference in median between high and low using original data
actual_median <- by(wb$Measles_change, wb$GNI_8000, median)
actual_median <- actual_median[1] - actual_median[2] #high - low

# permutations
diffmedians <- rep(NA, 10000)

for (i in 1:10000){
  s <- sample(wb$GNI_8000)
  diffmedians[i] <- median(wb$Measles_change[s == "high"]) - median(wb$Measles_change[s == "low"]) # original was high-low
}

# histogram
hist(diffmedians,
     main = "Histogram of Permuted Sample Medians \n of Difference in Measles Vaccination Rates \n between Countries based on GNI",
     xlab = "Measles Vaccination Rates",
     col = 'burlywood4',
     xlim = c(-8, 8))

# add vertical line at test hypothesis median
abline(v = actual_median,
       lwd = 3,
       col = 'peru')

text(3.75, 1650, paste("Actual Difference in Medians = ", actual_median), srt=90)

# obtain p value by counting what fraction of fake test statistics were more extreme than the actual one 
mean(abs(diffmedians) >= abs(actual_median))
```

*Null: there is no difference between median measles vaccination rate change between countries with a high and low GNI.*
*Alternative: there is a difference between median measles vaccination rate change between countries with a high and low GNI.*
*Futhermore, for permutation tests, the null hypothesis is that when switching the labels, the difference will be seen again, meaning that level of change can be seen by random chance and thus the null hypothesis must be true. The alternative hypothesis is that when switching the labels, the test statistic from the original test will not be seen and thus must not have happened by chance and then we can claim that there is a difference between the groups.*
*pvalue = 0.0019 so with an alpha of .01, we can reject the null hypothesis and claim that there is an actual difference between group medians.*


4)  Pseudo-coding

A public health group in New Haven is studying patterns of childhood and respiratory illnesses over the past 25 years, focusing on the fall season from August through mid-December. The dataset includes weekly counts of hand, foot, and mouth disease and strep throat from the first full week of August through the last full week of December.  There is also information on the academic and break/holiday schedules for Greater New Haven colleges as well as the public school system. Vaccination rates for each year (not week) are also available. Write pseudocode describing how you would explore how yearly illness incidence changes over time, test whether rates differ before and after school starts, compare high- versus low-vaccination years, and assess whether the observed differences are statistically significant.

*Like any dataset, you would want to start by looking it over; analyzing the dim, str, names, and analyzing head and tail. I am assuming the dates and vaccination would be in the dataset somehow, but if that information is separate, you would want to analyze it individually. You will likely want to create a lubridate function loop to ensure the schedules and dates and years are all expressed the same. It would also be helpful to run some summary statistics: the mean amount of weekly cases of hand and foot and strep, what is the typical range of the calendar the schools are open, what is the average vaccination rate.*
*Once you get familiar with the data, you will need to aggregate the weekly illness counts into "yearly" values. You can do this by summing them in a variable stored outside the df (avg_illness_year  <- sum(illness counts for that year)) or you could begin a new dataframe that is year, yearly counts of hfm, yearly counts of strep. Once you have this, you can use it to analyze the illness counts over the years. It might be helpful to make visualizations, like a histogram of the frequency of yearly illness counts or boxplots of the yearly illness counts for each illness(if the values are small, it may be necessary to do log(yearly illness) but by doing year this will be unlikely). Using the newly created variables or df, you will want to create a change in value variable that takes the yearly illness count from one year and subtracts the yearly illness count from the year prior. You could probably do this efficiently using a for loop that cycles the the pairs of years and stores the outcome in a column in the dataframe. You could then create a histogram to understand the frequency of the changes between years and the distribution of the overall changes between years. You could run a one sample t test on yearly change values to determine if there was a significant difference compared to no change (0). You could also execute a bootstrapping of the mean, median, or variance of illness change or yearly illness if we wanted to confirm any further tests of significant. This would likely be beneficial because the sample size of 25 is still relatively small. You would do this for each disease separately.*
*To investigate whether illness differs before and after school start, you can add a tag/variable that indicates if the week reported was in school or out of school in the original df. You can do this by analyzing the schedules and writing if then statements that will tag it as this school was in session or not (ex: if you look at the public school schedule make a "if this date is after (start date) and before (first break) then it would have DURING in the tag column"). You can create histograms and boxplots of the before, during, and after weekly illness counts to see the frequency and distribution of the data. You can then run t tests to compare the weekly illness counts before, after, and during school to determine if there is a significant difference (you could do this as multiple t tests (1v1) or an ANOVA (comparing all groups)). Finally, if you want to investigate further, you can do permutation test where you switch up the labels and see if there is still a difference between the weekly illness counts regardless of if they're before, during, or after school.*
*To analyze high vs low vaccination years, you first need to determine a cutoff of what is high and what is low. This will require some domain knowledge and consulting with an expert or the literature and will likely be specific to the vaccination (ex: we would expect/hope that measles vaccination rate might be higher than flu vaccination rate). You can create histograms and boxplots of the yearly vaccination rates to look at the distribution of the data. To compare illness in high vs low vaccination years, you could do a t test between the high and low groups. If you wanted to look further, you could also do a two sample t test between the averages of the first 12 years vs last 12 years to determine if there is a significant difference. Or you could do a one sample t test against a benchmark value, like the vaccination rate of the first reported year or the national average. If want to investigate further, you could do permutation test where you switch up the labels of "high" and "low" and see if there is a significant difference between the values. It might also be beneficial to determine a change in vaccination rates to compare to the change in illness per year value calculated earlier. You could add another variable to the newly created yearly df that is change in vaccination created similarly to the change in illness rate (vaccination year - vaccination year of year before). For the vaccination rate change, you could do a bootstrap to see if any characteristic (mean, median, variance) is significant. This would be beneficial because of the small sample size.*












