---
title: "Homework 03 Data Cleaning, Normal Distributions"
output:
  word_document: 
     fig_height: 5
     fig_width : 7
  pdf_document: 
     fig_height: 5
     fig_width : 7
  html_document: default
subtitle: Due by 11:59pm, Tuesday, 2.10.26
author: S&DS 2300/5300/ENV 757
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**(1) More on List Manipulation** *(18 points - 3 points each)*.

(1.1) Make an object called `myList` that contains the following
elements (in order):

-   A matrix with the integers 1 through 36 that has six rows, filled by
    row

-   A list that contains

    -   A vector with the text "Pianoforte" and "Gerbera Daisy"
    -   A vector with the integers 9 through 13

-   The integers 1 through 10

You should be able to make this object in a single line of code.

*Use `[]`, `[[]]`, `[,]` notation to answer parts b) through f).*

(1.2) Make an object called `ans1` that is the third row of the matrix
contained in `myList`.

(1.3) Make an object called `ans2` that is the sum of the 6th column of
the matrix contained in `myList`.

(1.4) Make an object called `ans3` that is the sum of EACH column of the
matrix contained in `myList` *(use the `apply()` function or check out
`colSums()`)*.

(1.5) Make an object called `ans4` that is the single element of
`myList` that you can play.

(1.6) Make an object called `ans5` that is the third element of the
second element of the second element of `myList` converted to
characters.

Get the results of each of your objects you created above (i.e. get them
to show up in your knitted file by typing their names or putting the
code line that creates each object in parentheses).

```{r}
# 1.1
(myList <- list(matrix(1:36, nrow = 6, byrow = TRUE), list(c("Pianoforte", "Gerbera Daisy"), c(9:13)), (1:10)))

# 1.2
(ans1 <- myList[[1]][3, ])

# 1.3
(ans2 <- sum(myList[[1]][6,]))

# 1.4
(ans3 <- colSums(myList[[1]]))

# 1.5 
(ans4 <- myList[[2]][[1]][1])

# 1.6
(ans5 <- as.character(myList[[2]][[2]][3]))

# check if it is a character - also returned with "" which indicates character, but just to be safe 
str(ans5)
```

**(2) Normal Quantile Plots and the Binomial Distribution** *(20 points,
3 points each, part (2.5) is 5 points)*.

You may recall from your Intro Statistics course that a binomial
distribution looks like a normal distribution if np \> 10 and n(1-p) \>
10 (i.e. as long as the average number of successes and failures are
both larger than 10). Recall that n is the number of trials, and p is
the probability of success for each Bernoulli trial. *As an example,
flip a coin 30 times, count the number of heads. n=30, p=.5, np = 15 \>
10 and n(1-p) = 15 \> 10, so the distribution should be approximately
normal)*.

You are going to make six normal quantile plots that simulate 127 random
observations from binomial distributions with p = 0.3 and various values
of n.

(2.1) Install the `car` package. This will allow you use the `qqPlot()`
function. Load this package.

(2.2) Make a vector called `vec` that is powers of 10 for powers 0
through 5. The one caveat is that you need to use the`**` operator which
reads as 'to the power of' *(i.e. `2**3` is 8).* Show what is contained
in `vec`.

(2.3) Use the `par()` function to set up your plot region to show 6
plots on a page. Go learn about the `mfrow` option in par to create a
plot that was two rows and three columns.

(2.4) Use the `rbinom()` function to generate 6 random binomial
observations, each with 17 trials, and with p=0.6. You may need to type
`?rbinom` to get the syntax for this function (it is somewhat
counterintuitive). Store the result in an object called `vec2` and show
what is contained in `vec2`.

(2.5) Write a loop that repeatedly creates a normal quantile plot for
127 random samples each from a binomial distribution with p = 0.3 and n
equal to the 6 values stored in `vec`. A few plot details: \* Use the
`qqPlot` function. \* Make the graph points red solid dots (`pch = 19`).
\* Make the boundary lines blue (use `col.lines`) \* Make a main graph
title that pastes the text "127 Binomial Samples, N =" to the
corresponding value from `vec`.

```{r}
# 2.1 install car package
# install.packages('car')
library('car')

# 2.2 make vector
vec <- 10 ** (0:5)
vec

# 2.3 use par() to set up plot
par(mfrow = c(2,3))

# 2.4 rbinom()
vec2 <- rbinom(6, 17, 0.6)
vec2

# 2.5 loop plotting
for (i in vec) {
  qqPlot(rbinom(127, i, 0.3),
  col = 'red',
  pch = 19,
  col.lines = "blue",
  main = paste("127 Binomial Samples, N = ", i)
  )
}

```

(2.6) Take a look at the normal quantile plots. For what value of n do
the graphs seem to be approximately normally distributed? Is this
consistent with what you expect? Write two complete sentences to answer
these questions.

*Graphs with 1,000, 10,000, and 100,000 trials looks approximately normal (the 100 trial size is nearing normal but still has some skew, especially compared to the other graphs that almost completely stay within the confidence interval bounds). This is consistent with what I expected because as the trial number increases, the more approximately normal and within the confidence interval the sample will look, following the central limit theorem.*

**(3) Favorite food and Data Cleaning** *(62 points. Parts 3.2 through
3.5, 2 pts each, other values listed below)*.

This is data generated by former students. I simply asked "What is your
favorite food?". You can get the data
[HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/Food_230.csv).

Your goal is similar to what we did with the question "What animal would
you like to be?" in Class 5: clean this variable, make a barplot, and
discuss the results.

(3.1) *(1 pt)* Read in the data to a new object called `food`.

(3.2) Get a sense of the dataset - dimensions, variable names, look at
the first few rows.

(3.3) Convert `food` to a single vector that is just the first column
(literally, replace `food` with `food$Food`). if you need to, convert
this value to character.

(3.4) Show the sorted unique values of `food`. Calculate how many unique
values exist in `food`.

```{r}
# 3.1
# read in data as csv
food <- read.csv("https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/Food_230.csv")

# 3.2
# dimensions of data - 317 rows/responses, 1 column/question
dim(food)

# names of columns - only column is named food
names(food)

# head and tail of data
head(food)
tail(food)

# structure of data - df of characters
str(food)

# 3.3 
# convert food to be single vector of the food column; confirmed was character string
food <- food$Food
str(food)

# 3.4 
# number of unique values - there are 218 
unique(food)
length(unique(food)) # display

# sort unique values
(sort(unique(food))) # display
```

(3.5) Write a couple of sentences about what data cleaning issues you
notice amoung the unique values of `food`.

*There are casing inconsistencies (burritos vs Burritos). Some people have additionally commentary (Thai food vs Thai food is my favorite). There are punctuation or random unknown character (\n in ramen, !!behind pizza). Misspellings could also be an issue. Some people did cuisines and some people did specific foods (Mexican food vs burritos) - this will require me to make a decision about how to handle and categorize these.*

(3.6) Cleaning Part I *(8 pts)*: Clean the data using the following
steps (in order):

-   Convert data to lower case
-   Find " or " and remove this and anything that follows.
-   Find " and " and remove this and anything that follows.
-   Find " food" and remove this AND anything that follows.
-   Find " cuisine" and remove this AND anything that follows.
-   Remove all special characters and punctuation - see class 6.
-   Remove trailing spaces at the end of text

At each step, you'll probably want to check what unique values of `food`
are left to make sure your functions are working correctly. By the time
you finish, you should have 156 unique levels.

Your final two lines of code should again show the sorted unique values
of food and the current number of unique values.

```{r}
# create a backup variable to go back to if needed
food_backup <- food

# convert data to lowercase
food <- tolower(food)
length(unique(food)) # down to 188

# find 'or' and remove anything that follow
food <- gsub(" or.*", "", food) # make sure to have spaces so or on it's own, not in middle of word
length(unique(food)) # still 188

# find 'and' and remove anything that follows
food <- gsub(" and.*", "", food)
length(unique(food)) # 187

# find 'food' and remove anything that follows 
food <- gsub(" food.*", "", food)
length(unique(food)) # 171

# find 'cuisine' and remove anything that follows
food <- gsub(" cuisine.*", "", food)
length(unique(food)) # 170

# remove all special characters and punctuation
food <- gsub("[^A-Za-z ]", "", food) # anything not A-Z or a-z or a space is dropped
length(unique(food)) # 157

# remove trailing spaces 
food <- gsub("\\s$", "", food)
food <- trimws(food)
length(unique(food)) # 156

# rerun sorted
sort(unique(food))
```

(3.7) Cleaning Part II *(10 pts)*: A few quick random cleaning items:

Clean up the following types of food (in order) - one line of code per
type of food. In each case, deal with misspellings, modifiers ("shrimp
curry" vs just "curry"), two words ('hot pot' instead of 'hotpot'),
plurals, etc.

-   hotpot
-   curry
-   lasagna
-   noodles
-   cookies
-   chocolate
-   cheese
-   steak
-   sushi
-   fries (french, cajun, five guys' should all be 'fries')
-   ramen
-   tofu
-   burgers (of any kind)
-   soup
-   anything containing 'delicious' just call 'delicious'

When you're finished, you should have 130 unique values.

Your final two lines of code should again show the sorted unique values
of food and the current number of unique values.

```{r}
# hotpot
food <- gsub(".*hotpot.*|.*hot pot.*", "hotpot", food)
length(unique(food)) #153

# curry
food <- gsub(".*curry.*|.*curries.*", "curry", food)
length(unique(food)) #149

# lasagna
food <- gsub("lasagne", "lasagna", food)
length(unique(food)) #148

# noodles
food <- gsub(".*noodles.*|.*noodle.*", "noodles", food)
length(unique(food)) #145

# cookies
food <- gsub(".*cookies.*", "cookies", food)
length(unique(food)) #144

# chocolate
food <- gsub(".*chocolate.*", "chocolate", food)
length(unique(food)) #142

# cheese
food <- gsub(".*cheese.*", "cheese", food)
length(unique(food)) #139

# steak
food <- gsub(".*steak.*", "steak", food)
length(unique(food)) #140

# sushi
food <- gsub(".*sushi.*", "sushi", food)
length(unique(food)) #137

# fries (french, cajun, five guys' should all be 'fries')
food <- gsub(".*fries.*", "fries", food)
length(unique(food)) #135

# ramen
food <- gsub(".*ramen.*", "ramen", food)
length(unique(food)) #134

# tofu
food <- gsub(".*tofu.*", "tofu", food)
length(unique(food)) #133

# burgers (of any kind)
food <- gsub(".*burgers.*", "burgers", food)
length(unique(food)) #132

# soup
food <- gsub(".*soup.*", "soup", food)
length(unique(food)) #131

# anything containing 'delicious' just call 'delicious'
food <- gsub(".*delicious.*", "delicious", food)
length(unique(food)) #130

sort(unique(food))
```

(3.8) Cleaning Part III *(8 pts)*: Cleaning types of cuisine.

Clean up the following types of cuisine (in order) - in this case,
you'll want to make a vector called `searchvec` that contains the types
of cuisine. Then create a loop following the example in Class 5 to
replace all the modifiers for each cuisine type so that you ultimately
end up with cleaned up versions of each cuisine type. Use not more than
5 lines of code.

The cuisine types (in order) are \* asian \* chinese \* vietnamese \*
italian \* indian \* thai \* mexican \* brazilian \* korean

(there are other types of cuisine, but they don't require cleaning).

When you're finished, you should have 120 unique values.

Your final two lines of code should again show the sorted unique values
of food and the current number of unique values.

```{r}
# create backup 2
food_backup2 <- food

# search vector of all cuisines
searchvec <- c("asian", "chinese", "vietnamese", "italian", "indian", "thai", "mexican", "brazilian", "korean")

# loop to replace all modifiers 
for (i in 1:length(searchvec)){
  food <- gsub(paste0(".*", searchvec[i], ".*"), searchvec[i], food)
}

# look at uniques 
sort(unique(food))
length(unique(food))

```

(3.9) *(15 pts)* Following the example from Class 05, display a
dataframe of the sorted tabular results of `food` to see how many
individuals prefer each kind of food.

From here on, the decisions of how to clean and combine categories are
yours! Any food that currently has a count of 3 or more should remain
(you can add to these categories - for example, you could add 'lasagna'
to 'italian' or to 'pasta'). All other levels should be recoded or
incorporated into a 'miscellaneous' food category. Points awarded based
on thoughtfulness, effort, and quality/preciseness of your code.

Include your code below, and add comments where appropriate to describe
the choices you make. You should have no more than 40 levels by the time
you finish.

Display a dataframe of the sorted tabular results of `food` to see how
many individuals prefer each kind of food AGAIN after you've finished
your coding.

```{r}
# original df 
(df_food <- data.frame(table(food))) 
sort(table(food), decreasing = TRUE, value = TRUE)# currently 120 levels 

# I noticed in the table there was a "my favorite kind of ..." I went to my backup and found pastries had been erranously removed behind food so I reinserted that. I also confirmed in the backup that freeze dried was the full entry and it was 
food <- gsub("my favorite kind of", "pastries", food)

# i am making a vegetable category to group together broader vegetables
food <- gsub("corn|sundried tomatoes|sundried tomatoes|sweet potato|fried okra|artichoke", "vegetables", food)

# i am adding to the fruit category to group together broader fruits
food <- gsub("nectarine|strawberries|bananas|mango|dried fruits|platanos|dried mangos|fruits", "fruit", food)

# i am making a breakfast category - must come before desserts bc of pancakes and handles a misspelling of breakfast foodss
food <- gsub("crepes|pastries|cinnamon roll|bagel|multigrain pancake|breakfast foodss", "breakfast foods", food) # must change pancakes before desserts

# i am making a desserts category to group together broader desserts - i decided against chocolate being a dessert on it's own because these other items could contain chocolate as an ingrediant therefore chocolate is more of an ingrediant
food <- gsub("cookies|cake|donuts|dessert|desserts", "desserts", food)

# other country dishes - must come before chicken dishes. These are dishes are are more hallmarks of other countries/nationalities but are not the whole cuisine and more than just an item 
food <- gsub("kelewele its an african dish|palak paneer|shahi paneer|tagine|farofa|jollof rice|hibachi|swedish meatballs|roast dinner|jollof rice with chicken dishes", "other national dishes", food)

# i am making a chicken dishes category - I debated a lot what to do with chicken tikka but because it is ultimately chicken and only the chicken was mentioned rather than a dish or a national cuisine, I left it here
food <- gsub("chicken|chicken wings|fried chicken|butter chicken|chicken malai kabab|chicken tikka masala|chicken parmesan|chicken tikka|chicken tenders|gizzard chicken|grilled chicken breast", "chicken dishes", food)

# i am making a snacks category to encompass broader snacks. This ends up being in other anyways, so potentially unnecessary
food <- gsub("cheez its|sweet chili doritos", "snacks", food)

# i am making a seafood category to encompass broader seafood dishes 
food <- gsub("salmon|blue point oyster|ceviche|fish tacos|fried fish|fish|poke bowl", "seafood", food)

# other country cuisines - any other country cuisines outright mentioned that received less than 3 votes
food <- gsub("brazilian|albanian|ethiopian|vietnamese|indonesian|lebanese|persian|puerto rican|mediterranean", "other national cuisines", food)

# specific pizza - adding these to pizza
food <- gsub("good pizza|lahmacun turkish pizza", "pizza", food)

# specific breads - adding these to bread
food <- gsub("arepa|baguettes", "bread", food) 

# specific meats -  adding these to meat
food <- gsub("ribs|ham", "meat", food) 

# wrapped things - once I got down to the bottom, there were many food items that I would considered wrapped and have a similar vibe 
food <- gsub("empanadas|enchiladas|dumplings|flautas|potstickers|gyros|postickers", "wrapped delicacies", food) 

# replacing entries under 3 to be other - I wanted anything under 3 to be other. Anything that initially started as a category with 3 or more votes  was not integrated into any other categories and therefore I wanted them to continue to be represented outside of other. 
freq <- table(food)
food[food %in% names(freq[freq < 3])] <- "Other"

length(unique(food)) # ended with 37 unique items 
sort(table(food), decreasing = TRUE, value = TRUE)

```

(3.10) *(8 pts)* Final steps and a plot: You'll want to CAREFULLY follow
the example in the code at the end of Class 05.

-   Use the `toTitleCase()` function from the package `tools` to convert
    food to title case.
-   Make an object called `finaltab` that is a table of your final
    vector `food`.
-   Calculate percents, rounded to the nearest integer, for each food
    type. Save this as an object called `percents`.
-   Change the names of `finaltab` to include a space and then the
    percents followed by a "%" in curved parentheses.
-   Make a horizontal barplot of your final plot. Choose a nice bar
    color, adjust the left margins as necessary, give a main title and
    label the horizontal axis.

```{r}
# install tools package for title case
# install.packages('tools')
library('tools')

# apply title case
food <- toTitleCase(food)

# make object called finaltab
finaltab <- table(food)

# calculate percents, rounded for each food 
percents <- round(finaltab/sum(finaltab)*100)

# change names
names(finaltab) <- paste0(names(finaltab)," (", percents, "%)")

# plot 
png("foods.png", width = 800, height = 1200) # makes it a photo; made it less fuzzy, all labels seen and not weirdly off-centered
par(mar = c(15, 15, 10, 2)) # change margin and text size 

barplot(sort(finaltab, decreasing = TRUE),  # puts biggest values on bottom
        col = "plum", 
        main = "Favorite Food of Class", 
        horiz = TRUE, 
        las = 1, 
        xlab = "Number of Responses Received", 
        xlim = c(0,30)
)
```

(3.11) *(3 pts)* In no more than three sentences, discuss your process and results.  Be sure to mention how many unique values of 'food' you started and ended with.  Any surprises?

*I started with 120 unique values and started by just visually analyzing the values to determine if there were any very similar things I could start grouping, like vegetables and fruits. After, those initial ones, I found it very tough to create groupings; I struggled to determine if something should be included in a cuisine (is a burrito Mexican?) but decided that because anything over 3 should remain, some cuisine and items would remain (this would leave both sushi and Japanese for example) not to stress too much about it and I created two large grouping of other national dishes and cuisines. I made everything less than a 3 be considered Other and this caused me to end with 37 unique values. *

(4.1)  Pseudocode.

You want to understand how to better promote tourism in New Haven and coastal Connecticut. A company collected open-ended survey responses to the question: “What is the main reason you would visit this area?” Each row is one person’s answer. The responses are messy: inconsistent capitalization, extra punctuation, multiple-word answers (“the beach!!!”), and people mixing categories (e.g., “pizza or seafood”). You want to: 
- Clean the responses so categories are consistent (lower case, remove punctuation, trim spaces)
- Group answers into broader themes (e.g., pizza → food, museums → culture, beach → outdoor recreation)
- Summarize the cleaned results into counts and percentages. 
- Visualize the final categories in a clear barplot.
- Write pseudocode describing the steps you would take. Don't write any R code, but you're welcome to mention functions you would use.

*First, I would make sure I understand my data by investigating the dim, str, and names of the columns. I would put all the responses in a single vector so they are easy to manipulate. I would save this single vector into a backup variable in case I need to check or restore any of the original values. I would then sort all unique values and get the length to understand how many unique values we are starting with. If it is manageable file, I may visually look at the unique values as well. First, I would lowercase all the words and remove any punctuation. After each step I would check the length of the unique values. I would also trim any white spaces, delete trailing spaces, or special characters that may be due to responses being submitted from different computers. If there are any common leading phrases, like "I love the..." I would drop these and keep only any characters that follow. I would also drop 'the' and 'a'. If there are an overwhelming amount of responses, I could drop 'or,' 'and,' and 'because' and everything after although I would personally prefer to not and determine this on a case by case basis. If there are any common misspellings (ex: baech instead of beach), I would fix these using gsub. If there are any common themes that I can begin to group in, like multiple parks or beaches, I can do gsub to have these responses only be reflective of the broader category. If there are multiple themes, I could potentially fix these using a loop and "or" (|) indicators that looks for the theme, and replaces it with that themes word. Hopefully at this point there are minimal enough themes that I can begin grouping them manually based on my rules and thoughts. The amount of rules and groupings to be applied would depend on how many end categories are needed or applicable. After this I can summarize the final results into a table, which would show counts, or a barplot which would show percentages. To create final percentages, I would have a value be divided by the sum of all values (finaltab/sum(finaltab)and multiplied by 100. This could be rounded to the closest integer or a decimal; they could also be displayed in the table, in the barplot, or in the barplot labels. To visualize the results in a barplot, it would likely be best to do a horizontal plot to best appreciate the differences. If the side labels are long, I might need to extend the left margin or adjust the character size to accomodate for those. I would want to ensure the casing of the labels and inclusion of any wanted percentages (using pasgte) are also considered prior to plotting. I would also ensure all labels, axis levels, and titles are clear to best encompass and show off the survey results and the other cleaning done. *