---
title: "Homework 07 Multiple Linear Regression"
output:
  word_document: default
  html_document: default
  pdf_document: default
subtitle: Due by 11:59pm, Tuesday, 4.7.26
author: S&DS 2300/5300/ENV 757
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Submission Details

This homework assignment contains 2 problems. Edit this `.Rmd` file and insert your responses under the appropriate problems. Submit the knitted output of this file as a .pdf to Gradescope. Be sure to match pages in your submitted PDF to the requested subquestions on Gradescope.

For full credit, your document should look nicely formatted, with R code, plots, and text descriptions nicely integrated.  

**BE SURE TO REMOVE EVERY INSTANCE OF `eval = F`**

**Please DO NOT put your name anywhere on your assignment.**

**DELETE ALL LINES FROM HERE UP THROUGH 'Submission Details' BEFORE KNITTING AND SUBMITTING!**

**1)  Model Overfitting and R-squared** 

In class, I discussed how, despite its merits, R-squared is not useful as the sole measure of predictive accuracy. The problem is that as more predictors are added (even useless ones), R-squared will always increase (assuming there is no missing data).

This problem is magnified in the case of **Model Overfitting**.   We say that a model is over-fit if the number of predictors approaches the number of observations.  When the number of predictors is one less than the number of observations, the R-squared will always be 1. 

We do a simulation to see this in action.

First, I'll simulate all of my variables (predictors and response) from a normal(0, 1) distribution. That is, the y values and ALL of the possible X predictor values are all randomly chosen from a random normal distribution with mean zero and sd = 1.  This means that the x variables are uncorrelated AND they are all NOT significant predictors of Y.


```{r}
set.seed(1)
simdata <- rnorm(10 * 15) # need 10 * 15 values simulated
simdata <- matrix(simdata, nrow = 10, ncol = 15) # now convert this vector into a matrix
colnames(simdata) <- c("y", paste0("x", 1:14)) # add column names
simdata <- as.data.frame(simdata) # convert the matrix into a data frame
```

Let's pretend that this is an actual dataset we'd like to use for linear regression, 10 observations (rows) of 14 predictors and a response variable. The columns are named as follows (so we treat the first column "y" as our response variable and the subsequent columns "x1", "x2", ... as our predictors):

```{r}
colnames(simdata)
```

**Note**: By virtue of how we simulated the data (independently and identically sampled from a standard normal distribution), "y" has no relationship with any of the 14 possible predictors.

1.1) Fit a simple linear regression model predicting `y` using `x1` and save the results to an object called `mod1`. Does it appear that `x1` is a significant predictor of `y`?  What is the value of R-squared (use `summary(mod1)$r.squared`)?  Interpret its value in the context of the model. 
 
```{r}
# fit simple linear regression model predicting y using x1
mod1 <- lm(y ~ x1, data = simdata)
summary(mod1$r.squared)

```


*Does it appear that `x1` is a significant predictor of `y`?*
*What is the value of R-squared (use `summary(mod1)$r.squared`)?*
*Interpret its value in the context of the model*

1.2) Using a for-loop, now expand your linear regression model in (a) by iteratively adding in one more predictor at a time. That is to say, the first iteration of your for-loop should use `x1` as a predictor; the second iteration of your for-loop should use `x1`, and `x2`; and so on, until all 14 predictors are used in your model. Store the values of R-squared in a vector called `rsqvals` of length 14, so that the `rsqvals[i]` should contain the value of R-squared for a model using predictors `x1` through `xi`.  Finally, display the values in `rsqvals`.

*Hint:* remember the shorthand formula to include all predictors: `y ~ .`. So, for example, if I wanted to fit a model using `x1` through `x7` as predictors, I could do:

```{r}
simtemp <- simdata[,1:8]
m7 <- lm(y ~ ., data = simtemp)
```


*REPLACE THIS TEXT WITH YOUR R-chunk Code*

1.3 How many predictors did it take to reach an R-squared of 1 (write a line of code to get this value)? Display a plot that shows the increase in R-squared with an increasing number of predictors.  Be sure to label your plot.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*


**2) Ohio Crime Data** 

A survey sponsored by the US Justice Department and the University of Cincinnatti interviewed a number of Ohio residents on their attitudes toward crime, criminals, and ways of reducing crime.  In addition, various religious and demographic information was collected.

A summary of survey questions (and label information) can be found [HERE](https://github.com/jreuning/sds230_data/blob/main/OhioCrime.pdf).  You'll want to look at pages 17 - 34 at a minimum.   Each question is labeled V1, V2, etc. through V98.   We'll be looking at a subset of these questions.

The data itself is [HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/ohiocrime.csv).

2.1)  Read the data into an object called `crime`.  Get the dimension and column names of `crime`.  You won't need the `as.is = TRUE` option.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.2)  First consider the variables in columns 10 through 23; these are 14 questions having to do with attitudes toward preventing crime (see PDF file).   Each question is on a 6 point scale (see PDF file for particular levels).  Your first task is to visually examine the correlations with the `corrplot.mixed` function discussed in class 11.

I've given an outline chunk below. Your job is to fill in the details as indicated.  Be sure to remove `eval = F` before knitting.


```{r fig.height = 5, fig.width = 5, eval = F}

#note the options above are to make plots work properly in the corrplot package.

#Load the corrplot package

cor1 <- #calculate pairwise correlations for columns 10-23 of crime.  You'll need the   use = "pairwise.complete.obs" option.

#round cor1 to 2 decimal places and display the result.  
  
#WRITE A DESCRIPTION HERE OF THE PURPOSE OF THE FOLLOWING LINE.
maxloc <- which(cor1 == max(cor1[cor1<1]), arr.ind = TRUE)

#WRITE A DESCRIPTION HERE OF THE PURPOSE OF THE FOLLOWING LINE.
names(crime[10:23])[maxloc[1,]]

#Create an object called sigcorr that has the results of cor.mtest for columns 10-23 of the crime data.  Use 95% CI.

#Use corrplot.mixed to display confidence ellipses, pairwise correlation values, and put on 'X' over non-significant values.

```


2.3)  Comment on the overall level of correlations among the considered questions.  Which pair of questions had the highest sample correlation?  Are you surprised? (include comments on the actual questions in your answer)

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

2.4) Make a scatterplot of values for the two questions that had the highest pairwise correlation.  Make sure your plot has labels for each axis (and not 'V10' - something with meaning).  Include two top titles  - one for the plot as a whole, one which reports the sample correlation to two decimal places.  How helpful is this plot?

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

2.5)  Repeat part 2.4) but jitter results in both directions.  Write a sentence about what you observe.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

We are now going to proceed with performing stepwise regression. In particular, we're going to fit a model that looks at possible predictors of question V45 (you'll want to look up what this question is).   To do this, I'm making a new dataset called `crime2` which contains the relevant columns (notice I'm putting the response variable FIRST).  Be sure to remove the option `eval = F`.

```{r, eval = F}
crime2 <- crime[, c(45, 10:23, 65, 70, 72, 87, 86)]
names(crime2)
dim(crime2)
```

2.6)  Perform best subsets regression using the `regsubsets` function in the `leaps` package.  Save the results in an object called `mod2`.  Get the summary of `mod2` and save the results in an object called `mod2sum`.  Display `mod2sum$which` to get a sense of which variables are included at each step of best subsets.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.7)  Following the example in classes 13-16, examine the best model according to highest r-squared.  Here are your steps:

*  Make an object called `modnum` which contains the row number in `mod2sum$which` for the model with the highest r-squared.
*  Print the variable names for predictors that ended up in this model.
*  Make a temporary dataset called `crimetemp` which has the columns of `crime2` that were included in this model.
*  Fit the model and return summary information for the model.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.8)  Repeat 2.7) for adjusted R-squared.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.9)  Repeat 2.7) for BIC.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.10)  Repeat 2.7) for the Cp Statistic.

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

2.11)  We choose as our final model the model indicated by BIC.  Go refit this model and save the results in an object called `modfin`.  How many observations had missing values in this model vs. the number of observations with missing values in the model with all predictors?

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

2.12)  Make two residual plots - studentized residuals vs fitted values (with boundaries at +/-2 and +/-3) as well a normal quantile plot of the residuals. Write a few sentences about how the plots do/do not indicate that we've met the assumptions of our regression model.  Note : the six-toed beast that seems to have slashed the plot of fits vs. residuals is exactly what we would expect.   Why is this?

*REPLACE THIS TEXT WITH YOUR R-chunk Code*

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

2.13)  Run the code below.  Describe what this does in not more than two sentences.

```{r, eval = F}

CIs <- confint(modfin)
coefs <- coef(modfin)
library(plotrix)
plotCI(1:(length(coefs)-1), coefs[-1], 
       ui = CIs[-1,2], li = CIs[-1,1], 
       xlab = "", 
       ylab = "Coefficient (and 95% CI)", 
       main = "Final Crime Model Coefficients and CI's", 
       axes = FALSE, lwd = 2, col = "blue")
abline(h = 0, lty = 2, lwd = 3, col = "red")
axis(side = 2)
axis(side = 1, at = 1:(length(coefs)-1), label = names(coefs)[-1], las = 2)

```

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

2.14)  FINALLY - write a short paragraph discussing your model results.

*  Comment on R-squared
*  Comment on direction and interpretation of each of the predictors included in the final model.   Are you surprised in any instance?

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

3)  Pseudocode

Prompt:
You’ve been hired by the Connecticut Department of Environmental Protection to help understand what drives nutrient pollution in lakes across the state. The dataset you’re given includes measurements for total phosphorus concentration (your response variable) and nineteen potential predictors, including lake area, depth, surrounding land use, population density, road density, and several management practices. Your boss wants you to determine which combination of predictors best explains variation in phosphorus levels, while also checking that your final model meets the assumptions of linear regression. Write pseudocode describing the steps you would take to provide meaningful insights into how different variables contribute to better or worse water quality (based on total phosphorus concentration) and what these variables may suggest about future protection, clean up, and management decisions (i.e. outline the steps you'd use to fit a multiple linear regression model).

*REPLACE THIS TEXT WITH YOUR Answer - make sure it's in italics*

